# Natural Experiments {#NE}

Natural Experiments are situations due to the natural course of events that approximate the conditions of a randomized controlled trial.
In the economists' toolkit, we generally make a distinction between:

  1. Instrumental variables (IV), that rely on finding a plausibly exogeneous source of variation in treatment intake.
  2. Regression Discontinuity Designs (RDD), that exploit a discontinuity in the eligibility to the treatment.
  3. Difference In Differences (DID), that make use of the differential exposure of some groups to the treatment of interest over time.
  
```{remark}
The term *Natural Experiments* seems to be mostly used by economists.
It dates back to Haavelmo (1944)'s paper on the Probability Approach to Econometrics, where he makes a distinction between the experiments we'd like to make as social scientists and the experiments that Nature provides us with, that are in general a subset of the experiments we'd like to make.
This raises the question of our ability to **identify** the relationships of interest from the variation that is present in the data, a traditional problem in classical econometrics that has echoes in treatment effect estimation, where we also try to *identify* treatment effect parameters.
At the time of Haavelmo, and until the beginning of the 1990s, there was no real discussion of the plausibility of the *identifying assumptions* (or restrictions) required for identification of certain relations, outside of a discussion of their theoretical plausiblility.
With the credibility revolution brought about by Angrist (1990)'s paper and summarized in Angrist and Krueger (2001)'s review paper, the notion of natural experiment made a come back, with the idea that we might be able to look for specific set of events produced by Nature that more credibly identify a relationship of interest, *i.e.* that closely approximate true experimental conditions.
```

```{remark}
Outside of economics, Natural Experiments have also flourished, but without the term, and were compiled in the early textbook on research methods by Campbell (1966).
Both Difference In Differences and Regression Discontinuity Designs have been actually developed outside of economics, mostly in education research.
Instrumental Variables have had a separate history in economics and in genetics, were it is called the method of path coefficients. 
```

## Instrumental Variables

Instrumental Variables rely on finding a plausibly exogeneous source of variation in treatment intake.
In the simple case of a binary instrument, the identification and estimation parts are actually identical to Encouragements designs in RCTs, that we have already studied in Section \@ref(sec:design4).
As a consequence, unless we make very strong assumptions, an IV design is going to recover a Local Average Treatment Effect.
Our classical assumptions are going to show up again: Independence, Exclusion Restriction, Monotonicity.

```{remark}
Examples of Instrumental Variables are:
```

  - Distance to college or to school for studying the impact of college or school enrollement on education, earnings and other outcomes.
  - Random draft lottery number for investigating the impact of military experience on earnings and other outcomes.
  - Randomized encouragement to participate in order to study the impact of a program.

```{remark}
The crucial part of an IV design is to justify the credibility of the exclusion restriction and independence assumptions.
It is in general very difficult to justify these assumptions, especially the exclusion restriction assumption. 
In the examples above, one could argue that schools or colleges might be built where they are necessary, i.e. close to destitute populations, or, on the contrary, that they are built far from difficult neighbourhoods. 
As soon as distance to school becomes correlated with other determinants of schooling, such as parental income and education, the independence assumption is violated.

Even if school placement is truly independent of potential education and earnings outcomes at first, parents, by choosing where to live, will sort themselves such as the parents that pay more attention to education end up located closer to school.
As a consequence, the independence assumption might be violated again.

Even when the instrument is truly random, such as a draft lottery number, and thus the independence assumption seems fine, the instrument may directly affect the outcomes by other ways than the treatment of interest. 
For example, receiving a low draft lottery number makes one more likely to be drafted. 
In response, one might decide to increase their length of stay in college in order to use the waiver for the draft reserved for students.
If receiving a low draft lottery number increases the number of years of education, and in turn subsequent earnings, then the exclusion restriction assumption is violated.
```

In this section, I'm going to denote $Z_i$ a binary instrument that can either take value $0$ or $1$.
In general, we try to reserve the value $1$ for the instrument value that increases participation in the treatment of interest.
In our examples, that would be when for example, the distance to college is low, the draft lottery number is low, or someone receives an encouragement to enter a program.

### An example where Monotonicity does not hold

Since Monotonicity is going to play such a particular role, and since we have already explored this assumption a little in Chapter \@ref(RCT), I am going to use as an example a model where the Monotonicity assumption actually does not hold. 
It will, I hope, help us understand better the way Monotonicity works and how it interacts with the other assumptions.
The key component of the model that makes Monotonicity necessary is the fact that treatment effects are heterogeneous and correlated with participation in the treatment.
We'll see later that Monotonicity is unnecessary when treatment effects are orthogonal to take up.

```{example}
Let's see how we can generate a model without Monotonicity:
```

\begin{align*}
y_i^1 & = y_i^0+\bar{\alpha}+\theta\mu_i+\eta_i \\
y_i^0 & = \mu_i+\delta+U_i^0 \\
U_i^0 & = \rho U_i^B+\epsilon_i \\
y_i^B & =\mu_i+U_i^B \\
U_i^B & \sim\mathcal{N}(0,\sigma^2_{U}) \\
D_i   & = \uns{y_i^B+\kappa_i Z_i + V_i\leq\bar{y}} \\
\kappa_i & = 
\begin{cases}
-\bar{\kappa} & \text{ if } \xi_i = 1 \\
\underline{\kappa} & \text{ if } \xi_i = 0
\end{cases} \\
\xi & \sim\mathcal{B}(p_{\xi}) \\
V_i   & = \gamma(\mu_i-\bar{\mu}) + \omega_i \\
(\eta_i,\omega_i) & \sim\mathcal{N}(0,0,\sigma^2_{\eta},\sigma^2_{\omega},\rho_{\eta,\omega}) \\
Z_i   & \sim\mathcal{B}(p_Z) \\
Z_i   & \Ind (y_i^0,y_i^1,y_i^B,V_i) \\
\xi_i & \Ind (y_i^0,y_i^1,y_i^B,V_i,Z_i)
\end{align*}

The key component of the model that generates a failure of Monotonicity is the coefficient $\kappa_i$, that determines how individuals' participation into the program reacts to the instrument $Z_i$.
$\kappa_i$ is a coefficient whose value varies accross the population.
In my simplified model, $\kappa_i$ can take only two values, $-\bar{\kappa}$ or $\underline{\kappa}$.
When $-\bar{\kappa}$ and $\underline{\kappa}$ have opposite signs (let's say $-\bar{\kappa}<0$ and $\underline{\kappa}>0$), then individuals with $\kappa_i=-\bar{\kappa}$ are going to be more likely to enter the program when they receive an encouragement (when $Z_i=1$) while individuals with $\kappa_i=\underline{\kappa}$ will be less likely to enter the program when $Z_i=1$.
When $-\bar{\kappa}$ and $\underline{\kappa}$ have different signs, we have four types of reactions when the instrumental variable moves from $Z_i=0$ to $Z_i=1$, holding everything else constant.
These four types of reactions define four types of individuals:

  * **Always takers** ($T_i=a$): individuals that participate in the program both when $Z_i=0$ and $Z_i=1$.
  * **Never takers** ($T_i=n$): individuals that do not participate in the program both when $Z_i=0$ and $Z_i=1$.
  * **Compliers** ($T_i=c$): individuals that do not participate in the program when $Z_i=0$ but that participate in the program when $Z_i=1$ .
  * **Defiers** ($T_i=d$): individuals that participate in the program when $Z_i=0$ but that do not participate in the program when $Z_i=1$ .
  
In our model, these four types are a function of $y_i^B+V_i$ and $\kappa_i$.
In order to see this let's define, as in Section \@ref(sec:design4), $D^z_i$ the participation decision of individual $i$ when the instrument is exogenously set to $Z_i=z$, with $z\in\left\{0,1\right\}$.
When $\kappa_i=-\bar{\kappa}<0$, we have three types of reactions to the instrument.
It turns out that each of type can be defined by where $y_i^B+V_i$ lies with respect to a series of thresholds:

  * **Always takers** ($T_i=a$) are such that $D^1_i=\uns{y_i^B-\bar{\kappa} + V_i\leq\bar{y}}=1$ and $D^0_i=\uns{y_i^B + V_i\leq\bar{y}}=1$, so that they actually are such that: $y_i^B+V_i\leq\bar{y}$.
  This is because $y_i^B+V_i\leq\bar{y} \Rightarrow y_i^B+V_i\leq\bar{y}+\bar{\kappa}$, when $\bar{\kappa}>0$.
  * **Never takers** ($T_i=n$) are such that $D^1_i=\uns{y_i^B-\bar{\kappa} + V_i\leq\bar{y}}=0$ and $D^0_i=\uns{y_i^B + V_i\leq\bar{y}}=0$, so that they actually are such that: $y_i^B+V_i>\bar{y}+\bar{\kappa}$.
  This is because $y_i^B+V_i>\bar{y}+\bar{\kappa} \Rightarrow y_i^B+V_i>\bar{y}$, when $\bar{\kappa}>0$.
  * **Compliers** ($T_i=c$) are such that $D^1_i=\uns{y_i^B-\bar{\kappa} + V_i\leq\bar{y}}=1$ and $D^0_i=\uns{y_i^B + V_i\leq\bar{y}}=0$, so that they actually are such that: $\bar{y}<y_i^B+V_i\leq\bar{y}+\bar{\kappa}$.
  
When $\kappa_i=\underline{\kappa}>0$, we have three types defined by where $V_i$ lies with respect to a series of thresholds:

  * **Always takers** ($T_i=a$) are such that $D^1_i=\uns{y_i^B+\underline{\kappa} + V_i\leq\bar{y}}=1$ and $D^0_i=\uns{y_i^B + V_i\leq\bar{y}}=1$, so that they actually are such that: $y_i^B+V_i\leq\bar{y}-\underline{\kappa}$.
  This is because $y_i^B+V_i\leq\bar{y}-\underline{\kappa} \Rightarrow y_i^B+V_i\leq\bar{y}$, when $\underline{\kappa}>0$.
  * **Never takers** ($T_i=n$) are such that $D^1_i=\uns{y_i^B-\bar{\kappa} + V_i\leq\bar{y}}=0$ and $D^0_i=\uns{y_i^B + V_i\leq\bar{y}}=0$, so that they actually are such that: $y_i^B+V_i>\bar{y}$.
  This is because $y_i^B+V_i>\bar{y} \Rightarrow y_i^B+V_i\leq\bar{y}-\underline{\kappa}$, when $\underline{\kappa}>0$.
  * **Defiers** ($T_i=d$) are such that $D^1_i=\uns{y_i^B+\underline{\kappa} + V_i\leq\bar{y}}=0$ and $D^0_i=\uns{y_i^B + V_i\leq\bar{y}}=1$, so that they actually are such that: $\bar{y}-\underline{\kappa}<V_i+y_i^B\leq\bar{y}$.

Let's visualize how this works in a plot.
Before that, let's generate some data according to this process.
For that, let's choose values for the new parameters.

```{r param.IV,eval=TRUE,echo=TRUE,results='hide'}
param <- c(8,.5,.28,1500,0.9,0.01,0.05,0.05,0.05,0.1,0.1,7.98,0.5,1,0.5,0.9,0.28,0)
names(param) <- c("barmu","sigma2mu","sigma2U","barY","rho","theta","sigma2epsilon","sigma2eta","delta","baralpha","gamma","baryB","pZ","barkappa","underbarkappa","pxi","sigma2omega","rhoetaomega")
```

```{r simul.IV,eval=TRUE,echo=TRUE,results='hide'}
set.seed(1234)
N <-1000
cov.eta.omega <- matrix(c(param["sigma2eta"],param["rhoetaomega"]*sqrt(param["sigma2eta"]*param["sigma2omega"]),param["rhoetaomega"]*sqrt(param["sigma2eta"]*param["sigma2omega"]),param["sigma2omega"]),ncol=2,nrow=2)
eta.omega <- as.data.frame(mvrnorm(N,c(0,0),cov.eta.omega))
colnames(eta.omega) <- c('eta','omega')
mu <- rnorm(N,param["barmu"],sqrt(param["sigma2mu"]))
UB <- rnorm(N,0,sqrt(param["sigma2U"]))
yB <- mu + UB 
YB <- exp(yB)
Ds <- rep(0,N)
Z <- rbinom(N,1,param["pZ"])
xi <- rbinom(N,1,param["pxi"]) 
kappa <- ifelse(xi==1,-param["barkappa"],param["underbarkappa"])
V <- param["gamma"]*(mu-param["barmu"])+eta.omega$omega
Ds[yB+kappa*Z+V<=log(param["barY"])] <- 1 
epsilon <- rnorm(N,0,sqrt(param["sigma2epsilon"]))
U0 <- param["rho"]*UB + epsilon
y0 <- mu +  U0 + param["delta"]
alpha <- param["baralpha"]+  param["theta"]*mu + eta.omega$eta
y1 <- y0+alpha
Y0 <- exp(y0)
Y1 <- exp(y1)
y <- y1*Ds+y0*(1-Ds)
Y <- Y1*Ds+Y0*(1-Ds)
```

We can now define the types variable $T_i$:

```{r types.illustration,eval=TRUE,echo=TRUE,results='hide'}
D1 <- ifelse(yB+kappa+V<=log(param["barY"]),1,0)
D0 <- ifelse(yB+V<=log(param["barY"]),1,0)
AT <- ifelse(D1==1 & D0==1,1,0)
NT <- ifelse(D1==0 & D0==0,1,0)
C <- ifelse(D1==1 & D0==0,1,0)
D <- ifelse(D1==0 & D0==1,1,0)
Type <- ifelse(AT==1,'a',
            ifelse(NT==1,'n',
                   ifelse(C==1,'c',
                          ifelse(D==1,'d',""))))

data.non.mono <- data.frame(cbind(Type,C,NT,AT,D1,D0,Y,y,Y1,Y0,y0,y1,yB,alpha,U0,eta.omega$eta,epsilon,Ds,kappa,xi,Z,mu,UB))
```

```{r plottypes,eval=TRUE,echo=TRUE,fig.cap='Types',fig.subcap=c('$\\kappa=\\bar{\\kappa}$','$\\kappa=\\underline{\\kappa}$'),fig.align='center',out.width='50%',fig.pos='htbp'}
#ggplot(data.non.mono, aes(x=V, y=yB),color(as.factor(Type))) +
#    geom_point(shape=1)+
#    facet_grid(.~ as.factor(kappa))

plot(yB[AT==1 & kappa==-param["barkappa"]]+V[AT==1 & kappa==-param["barkappa"]],y[AT==1 & kappa==-param["barkappa"]],pch=1,xlim=c(5,11),ylim=c(5,11),xlab='yB+V',ylab="Outcomes")
points(yB[NT==1 & kappa==-param["barkappa"]]+V[NT==1 & kappa==-param["barkappa"]],y[NT==1 & kappa==-param["barkappa"]],pch=1,col='blue')
points(yB[C==1 & kappa==-param["barkappa"]]+V[C==1 & kappa==-param["barkappa"]],y[C==1 & kappa==-param["barkappa"]],pch=1,col='red')
points(yB[D==1 & kappa==-param["barkappa"]]+V[D==1 & kappa==-param["barkappa"]],y[D==1 & kappa==-param["barkappa"]],pch=1,col='green')
abline(v=log(param["barY"]),col='red')
abline(v=log(param["barY"])+param['barkappa'],col='red')
#abline(v=log(param["barY"])-param['underbarkappa'],col='red')
text(x=c(log(param["barY"]),log(param["barY"])+param['barkappa']),y=c(5,5),labels=c(expression(bar('y')),expression(bar('y')+bar(kappa))),pos=c(2,4),col=c('red','red'),lty=c('solid','solid'))
legend(5,10.5,c('AT','NT','C','D'),pch=c(1,1,1,1),col=c('black','blue','red','green'),ncol=1)
title(expression(kappa=bar(kappa)))

plot(yB[AT==1 & kappa==param["underbarkappa"]]+V[AT==1 & kappa==param["underbarkappa"]],y[AT==1 & kappa==param["underbarkappa"]],pch=1,xlim=c(5,11),ylim=c(5,11),xlab='yB+V',ylab="Outcomes")
points(yB[NT==1 & kappa==param["underbarkappa"]]+V[NT==1 & kappa==param["underbarkappa"]],y[NT==1 & kappa==param["underbarkappa"]],pch=1,col='blue')
points(yB[C==1 & kappa==param["underbarkappa"]]+V[C==1 & kappa==param["underbarkappa"]],y[C==1 & kappa==param["underbarkappa"]],pch=1,col='red')
points(yB[D==1 & kappa==param["underbarkappa"]]+V[D==1 & kappa==param["underbarkappa"]],y[D==1 & kappa==param["underbarkappa"]],pch=1,col='green')
abline(v=log(param["barY"]),col='red')
#abline(v=log(param["barY"])-param['barkappa'],col='red')
abline(v=log(param["barY"])-param['underbarkappa'],col='red')
text(x=c(log(param["barY"]),log(param["barY"])-param['underbarkappa']),y=c(5,5),labels=c(expression(bar('y')),expression(bar('y')-underbar(kappa))),pos=c(2,2),col=c('red','red'),lty=c('solid','solid'))
legend(5,10.5,c('AT','NT','C','D'),pch=c(1,1,1,1),col=c('black','blue','red','green'),ncol=1)
title(expression(kappa=underbar(kappa)))

```

As Figure \@ref(fig:plottypes) shows how the different types interact with $\kappa_i$.
When $\kappa_i=-\bar{\kappa}$, individuals with $y_i^B+V_i$ below $\bar{y}$ always take the program.
Even when $Z_i=1$ and $\bar{\kappa}$ is subtracted from their index, it is still low enough so that they get to participate.
When $y_i^B+V_i$ is in between $\bar{y}$ and $\bar{y}+\bar{\kappa}$, the individuals are such that their index without subtracting $\bar{\kappa}$ is above $\bar{y}$, but it is below $\bar{y}$ when $\bar{\kappa}$ is subtracted from it.
These individuals participate when $Z_i=1$ and do not participate when $Z_i=0$: they are compliers.
Individuals such that $y_i^B+V_i$ is above $\bar{y}+\bar{\kappa}$ will have an index above $\bar{y}$ whether we substract $\bar{\kappa}$ from it or not.
They are never takers.

When $\kappa_i=\underline{\kappa}$, individuals with $y_i^B+V_i$ below $\bar{y}-\underline{\kappa}$ always take the program.
Even when $Z_i=0$ and $\underline{\kappa}$ is not subtracted from their index, it is still low enough so that they get to participate.
When $y_i^B+V_i$ is in between $\bar{y}-\underline{\kappa}$ and $\bar{y}$, the individuals are such that their index without adding $\underline{\kappa}$ is below $\bar{y}$, but it is above $\bar{y}$ when $\underline{\kappa}$ is added to it.
These individuals participate when $Z_i=0$ and do not participate when $Z_i=1$: they are defiers.
Individuals such that $y_i^B+V_i$ is above $\bar{y}$ will have an index above $\bar{y}$ whether we add $\underline{\kappa}$ from it or not.
They are never takers.

### Identification 

We need several assumptions for identification in an Instrumental Variable framework.
We are going to explore two sets of assumption that secure the identification of two different parameters: 

  * The Average Treatment Effect on the Treated ($TT$): identification will happen through the assumption of independence of treatment effects from potential treatment choice
  * The Local Average Treatment Effect ($LATE$)

```{hypothesis,FirstStage,name='First Stage Full Rank'}
We assume that the instrument $Z_i$ has a direct effect on treatment participation:

\begin{align*}
\Pr(D_i=1|Z_i=1)\neq\Pr(D_i=1|Z_i=0).
\end{align*}
```

```{example}
Let's see how this assumption works in our example.
Let's first compute the average values of $Y_i$ and $D_i$ as a function of $Z_i$, for later use.
```

```{r graph.illus.IV,eval=TRUE,echo=TRUE,results='markup'}
means.IV <- c(mean(Ds[Z==0]),mean(Ds[Z==1]),mean(y0[Z==0]),mean(y0[Z==1]),mean(y[Z==0]),mean(y[Z==1]),0,1)
means.IV <- matrix(means.IV,nrow=2,ncol=4,byrow=FALSE,dimnames=list(c('Z=0','Z=1'),c('D','y0','y','Z')))
means.IV <- as.data.frame(means.IV)
```

```{r IVFirstStage,eval=TRUE,echo=FALSE,results='hide',echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,fig.cap='Proportion of participants as a function of $Z_i$',fig.align='center',out.width=cst,fig.pos='htbp'}
ggplot(means.IV, aes(x=as.factor(Z), y=D)) +
  geom_bar(position=position_dodge(), stat="identity", colour='black')+
  xlab('Z')+
  ylab('Pr(D=1|Z)')
```

Figure \@ref(fig:IVFirstStage) shows that the proportion of treated when $Z_i=1$ in our sample is equal to `r round(mean(Ds[Z==1]),2)` while the proportion of treated when $Z_i=0$ is equal to  `r round(mean(Ds[Z==0]),2)`, in accordance with Assumption \@ref(hyp:FirstStage).
In the population, the proportion of treated when $Z_i=1$ depends on the value of $\kappa_i$. 
Let's derive its value:

\begin{align*}
  \Pr(D_i=1|Z_i=1) & = \Pr(y_i^B+\kappa_i Z_i + V_i\leq\bar{y}|Z_i=1) \\
                  & =  \Pr(y_i^B+\kappa_i + V_i\leq\bar{y}) \\
                  & =  \Pr(y_i^B+ V_i\leq\bar{y}+\bar{\kappa}|\xi_i=1)\Pr(\xi_i=1) + \Pr(y_i^B+V_i\leq\bar{y}-\underline{\kappa}|\xi_i=0)\Pr(\xi_i=0) \\
                                    & =  \Pr(y_i^B+ V_i\leq\bar{y}+\bar{\kappa})p_{\xi} + \Pr(y_i^B+ V_i\leq\bar{y}-\underline{\kappa})(1-p_{\xi}) \\
                  & =  p_{\xi}\Phi\left(\frac{\bar{y}+\bar{\kappa}-\bar{\mu}}{\sqrt{(1+\gamma^2)\sigma^2_{\mu}+\sigma^2_{U}+\sigma^2_{\omega}}}\right) + (1-p_{\xi})\Phi\left(\frac{\bar{y}-\underline{\kappa}-\bar{\mu}}{\sqrt{(1+\gamma^2)\sigma^2_{\mu}+\sigma^2_{U}+\sigma^2_{\omega}}}\right)
\end{align*}

where the second equality follows from $Z_i$ being independent of $(y_i^0,y_i^1,y_i^B,V_i)$, the third equality follows from $\xi_i$ being independent from  $(y_i^0,y_i^1,y_i^B,V_i,Z_i)$ and the last equality follows from the formula for the cumulative of a normal distribution. 
The formula for $\Pr(D_i=1|Z_i=0)$ is the same except for $\bar{\kappa}$ and $\underline{\kappa}$ that are set to zero.

Let's write two functions to compute these probabilities:

```{r prob.IV,eval=TRUE,echo=TRUE,results='hide'}
prob.D.Z.1 <- function(param){
  part.1 <- param['pxi']*pnorm((log(param["barY"])+param['barkappa']-param['barmu'])/sqrt((1+param['gamma']^2)*param['sigma2mu']+param["sigma2U"]+param['sigma2omega']))
  part.2 <- (1-param['pxi'])*pnorm((log(param["barY"])-param['underbarkappa']-param['barmu'])/sqrt((1+param['gamma']^2)*param['sigma2mu']+param["sigma2U"]+param['sigma2omega']))
  return(part.1+part.2)
}
prob.D.Z.0 <- function(param){
  part.1 <- param['pxi']*pnorm((log(param["barY"])-param['barmu'])/sqrt((1+param['gamma']^2)*param['sigma2mu']+param["sigma2U"]+param['sigma2omega']))
  part.2 <- (1-param['pxi'])*pnorm((log(param["barY"])-param['barmu'])/sqrt((1+param['gamma']^2)*param['sigma2mu']+param["sigma2U"]+param['sigma2omega']))
  return(part.1+part.2)
}
```

With these functions, we know that, in the population, $\Pr(D_i=1|Z_i=1)=$ `r round(prob.D.Z.1(param),2)` and $\Pr(D_i=1|Z_i=0)=$ `r round(prob.D.Z.0(param),2)`, which is not far from what we have found in our sample.

Our next set of assumptions imposes that the instrument has no direct effect on the outcome and that it is not correlated with all the potential outcomes.
Let's start with the exclusion restriction:

```{hypothesis,ExclusionRestriction,name='Exclusion Restriction'}
We assume that there is no direct effect of $Z_i$ on outcomes:

\begin{align*}
\forall d,z \in \left\{0,1\right\}\text{, } Y_i^{d,z} = Y_i^d.
\end{align*}
```

```{example}
In our example, this assumption is automatically satisfied.
```
Indeed, $y_i^{d,z}=y_i^0 + d(y_i^1-y_i^0)$ which is parameterized as $y_i^{d,z}=\mu_i+\delta+U_i^0+d(\bar{\alpha}+\theta\mu_i+\eta_i)$.
Since $y_i^{d,z}$ does not depend on $z$, we have $y_i^{d,z} = y_i^d$, $\forall d,z \in \left\{0,1\right\}$.
The assumption would not be satisfied if $Z_i$ entered the equations for $y_i^0$ or $y_i^1$.
For example, if $Z_i$ is the Vietnam draft lottery number (high or low) used by Angrist to study the impact of army experience on earnings, the exclusion restriction would not work if $Z_i$ was directly influencing outcomes, independent of miitary experience, by example by generating a higher education level. 
In that case, we could have $E_i=\alpha+\beta Z_i + v_i$, where $E_i$ is education, and, for example, $y_i^0=\mu_i+\delta+\lambda E_i+U_i^0$.
We then have $y_i^{d,z}=\mu_i+\delta +\lambda(\alpha+\beta z + v_i) +U_i^0+d(\bar{\alpha}+\theta\mu_i+\eta_i)$ which depends on $z$ and thus the exclusion restriction does not hold any more. 

Let us now state the independence assumption:

```{hypothesis,Independence,name='Independence'}
We assume that $Z_i$ is independent from the other determinants of $Y_i$ and $D_i$:

\begin{align*}
(Y_i^1,Y_i^0,D_i^1,D_i^0)\Ind Z_i.
\end{align*}
```

```{remark}
Why do we say that independence from the potential outcomes is the same as independence from the other determinants of $Y_i$ and $D_i$?
Because the only sources of variation that remain in $Y_i^d$ and $D_i^z$ are the other sources of variations (that is not the treatment $D_i=d$ nor the instrument variable $Z_i=z$).
```

```{example}
In our example, this assumption is also satisfied.
```
If we assumed that unobserved determinants of earnings contained in $U^0_i$ are correlated with the instrument value, then we would have a problem.
For example, if children that leave close to college have also rich parents, or parents that spend a lot of time with them, or parents with large networks, there probably is a correlation between distance to college and earnings in the absence of the program.
For the draft lottery example, you might have that people with a high draft lottery number who have well-connected parents obtain discharges on special medical grounds. 
Is that a violation of the independence assumption?
Actually no.
Indeed, these individuals are simply going to become never takers (they avoid the draft whatever their lottery number).
But $Z_i$ is still independent from the level of connections of the parents.
For the independence assumption to fail in the draft lottery number example, you would need that children of well-connected parents obtain lower lottery numbers because the lottery is rigged.
In that case, since well-connected individuals would have had higher earnings even absent the lottery, there is a negative correlation between $y_i^0$ and having a high draft lottery number ($Z_i$).  

The last assumption we need in order to identify the Local Average Treatment Effect is that of Monotonicity.
We already know this assumption:

```{hypothesis,MonotonicityIV,name='Monotonicity'}
We assume that the instrument moves everyone in the population in the same direction:

\begin{align*}
\forall i\text{, either } D^1_i\geq D_i^0 \text{ or } D^1_i\leq D_i^0.
\end{align*}
```

Without loss of generality, we generally assume that $\forall i$, $D^1_i\geq D_i^0$.
As a consequence, there are no defiers.

```{example}
In our example, this assumption is not satisfied.
```
There are defiers, as Figure \@ref(fig:plottypes) shows, when $\xi_i = 0$ and thus $\kappa_i=\underline{\kappa}$. 
Indeed, in that case, for the individuals who are such that $\bar{y}-\underline{\kappa}<y_i^B+V_i\leq\bar{y}$, we have $D^1_i=\uns{y_i^B+\underline{\kappa} + V_i\leq\bar{y}}=0$ and $D^0_i=\uns{y_i^B + V_i\leq\bar{y}}=1$.
This would happen for example if some people would go to college less if their house is located closer to the college, maybe for example because they have a preference not to stay at their parents' house. 

```{remark}
Why are defiers a problem for the instrumental variable strategy?
Because the Intention to Treat Effect that measures the difference in expected outcomes at the two levels of the instrument is going to be characterized by two-way flows in and out of the program, as we have already seen with Theorem \@ref(thm:ITELATE).
This means that some treatment effects will have negative weights in the ITE formula. 
In that case, you might have a negative Intention to Treat Effect despite the treatment having positive effects for everyone, or you might under estimate the true effect of the treatment.
This matters only when the treatment effects are heterogeneous. 
```

```{example}
Let us detail how non-monotonicity and the existence defiers act on the ITE in our example, since we now have defiers.
The first very important thing to understand is that all the problems we have happend because treatment effects are heterogeneous **AND** they are correlated with the type of individuals: defiers and compliers do not have the same distribution of treatment effects and, case in point, they do not have the same average treatment effects. 
The average effects of the treatment on compliers and defiers are not the same. 
Let us first look at the distribution of treatment effects among compliers and defiers in the sample and in the population.
```

In order to derive the distribution of $\alpha_i$ conditional on Type in the population, we need to derive the joint distribution of $\alpha_i$ and $y_i^B+V_i$ and use the **trmtvnorm** package to recover its density when it is truncated.
This distribution is normal and fully characterized by its mean and covariance matrix.

\begin{align*}
  (\alpha_i,y_i^B+V_i) & \sim \mathcal{N}\left(\bar{\alpha}+\theta\bar{\mu},\bar{\mu},
                                        \left(\begin{array}{cc}
                                              \theta^2\sigma^2_{\mu}+\sigma^2_{\eta} & (\theta+\gamma)\sigma^2_{\mu}+\rho_{\eta,\omega}\sigma^2_{\eta}\sigma^2_{\omega}\\
                                              (\theta+\gamma)\sigma^2_{\mu}+\rho_{\eta,\omega}\sigma^2_{\eta}\sigma^2_{\omega} &  (1+\gamma^2)\sigma^2_{\mu}+\sigma^2_U+\sigma^2_{\omega}\\
                                              \end{array}
                                        \right)
                                      \right)
\end{align*}

Let us write a function to generate them.

```{r AlphaTypesPopulation,eval=TRUE,echo=TRUE,results='hide'}
mean.alpha.yBV <- c(param['baralpha']+param['theta']*param['barmu'],param['barmu'])
cov.alpha.yBV <- matrix(c((param['theta']^2)*param['sigma2mu']+param['sigma2eta'],
                          (param['theta']+param['gamma'])*param['sigma2mu']+param['rhoetaomega']*param['sigma2eta']*param['sigma2omega'],
                            (param['theta']+param['gamma'])*param['sigma2mu']+param['rhoetaomega']*param['sigma2eta']*param['sigma2omega'],
                            (1+param['gamma']^2)*param['sigma2mu']+param['sigma2U']+param['sigma2omega']),2,2,byrow=TRUE)
# density of alpha for compliers
lower.cut.comp <- c(-Inf,log(param['barY']))
upper.cut.comp <- c(Inf,log(param['barY'])+param['barkappa'])
d.alpha.compliers <- function(x){
  return(dtmvnorm.marginal(xn=x,n=1,mean=mean.alpha.yBV,sigma=cov.alpha.yBV,lower=lower.cut.comp,upper=upper.cut.comp))
}
# density of alpha for defiers
lower.cut.def <- c(-Inf,log(param['barY']-param['underbarkappa']))
upper.cut.def <- c(Inf,log(param['barY']))
d.alpha.defiers <- function(x){
  return(dtmvnorm.marginal(xn=x,n=1,mean=mean.alpha.yBV,sigma=cov.alpha.yBV,lower=lower.cut.def,upper=upper.cut.def))
}
```

Let us now plot the empirical and theoretical distributions of the treatment effects for compliers and defiers.

```{r AlphaTypesPlot,eval=TRUE,echo=TRUE,fig.cap='Distribution of treatment effects by Type in the sample (dashed line) and in the population (full line)',fig.align='center',out.width='50%',fig.pos='htbp'}
# building the data frame
alpha.types <- as.data.frame(cbind(alpha,C,D,AT,NT)) %>%
                mutate(
                  Type = ifelse(AT==1,"Always Takers",
                                ifelse(NT==1,"Never Takers",
                                       ifelse(C==1,"Compliers","Defiers")))
                ) %>%
                mutate(Type = as.factor(Type))

ggplot(filter(alpha.types,Type=="Compliers" | Type=="Defiers"), aes(x=alpha, colour=Type)) + 
  geom_density(linetype="dashed") +
  geom_function(fun = d.alpha.compliers, colour = "red") +
  geom_function(fun = d.alpha.defiers, colour = "blue") +
  ylab('density') +
  theme_bw()
```

Figure \@ref(fig:AlphaTypesPlot) shows that the two distributions are actually very similar in our example. 
The distribution for the compliers is slightly above that for the defiers, meaning that the defiers should have lower expected outcomes in the population.
Let us check that by computing the average outcomes of compliers and defiers both in the sample and in the population.

```{r AlphaTypesMean,eval=TRUE,echo=TRUE,results='hide'}
# sample means
mean.alpha.compliers.samp <- mean(alpha[C==1]) 
mean.alpha.defiers.samp <- mean(alpha[D==1]) 

# population means
mean.alpha.compliers.pop <- mtmvnorm(mean=mean.alpha.yBV,sigma=cov.alpha.yBV,lower=lower.cut.comp,upper=upper.cut.comp,doComputeVariance=FALSE)[[1]]
mean.alpha.defiers.pop <-  mtmvnorm(mean=mean.alpha.yBV,sigma=cov.alpha.yBV,lower=lower.cut.def,upper=upper.cut.def,doComputeVariance=FALSE)[[1]]
```

In the population, the average treatment effect for compliers is equal to `r round(mean.alpha.compliers.pop[[1]],2)` and the average treatment effect for defiers is equal to `r round(mean.alpha.defiers.pop[[1]],2)`.
In the sample, the average treatment effect for compliers is equal to `r round(mean.alpha.compliers.samp[[1]],2)` and the average treatment effect for defiers is equal to `r round(mean.alpha.defiers.samp[[1]],2)`.

The difference between the treatment effect for compliers and defiers is a problem for the Wald estimator.
Let's look at how the Wald estimator behaves in the population (in order to avoid considerations due to sampling noise).
By Theorem \@ref(thm:ITELATE), the numerator of the Wald estimator is equal to the difference between the average treatment on compliers and the average treatment effect on defiers weighted by their respective proportions in the population. 
In order to be able to compute the Wald estimator, we need to compute the proportion of compliers and of defiers in the population. 
These proportions are equal to:

\begin{align*}
  \Pr(T_i=c) & = \Pr(\bar{y}< y_i^B+V_i\leq\bar{y}+\bar{\kappa}\cap\kappa_i=-\bar{\kappa}) \\
              & = \Pr(\bar{y}< y_i^B+V_i\leq\bar{y}+\bar{\kappa})p_{\xi} \\
  \Pr(T_i=d) & = \Pr(\bar{y}-\underline{\kappa}< y_i^B+V_i\leq\bar{y}\cap\kappa_i=\underline{\kappa}) \\
              & = \Pr(\bar{y}-\underline{\kappa}< y_i^B+V_i\leq\bar{y})(1-p_{\xi}),
\end{align*}

where the second equality follows from the fact that $\xi$ is independent from $y_i^B+V_i$ and uses the fact that $\Pr(A\cap B)=\Pr(A|B)\Pr(B)$.
Since $y_i^B+V_i$ is normally distributed and we know its mean and variance, these proportions can be computed as: 

\begin{align*}
  \Pr(T_i=c) & = p_{\xi}\left(\Phi\left(\frac{\bar{y}+\bar{\kappa}-\bar{\mu}}{\sqrt{(1+\gamma^2)\sigma^2_{\mu}+\sigma^2_U+\sigma^2_{\omega}}}\right)
                  -\Phi\left(\frac{\bar{y}-\bar{\mu}}{\sqrt{(1+\gamma^2)\sigma^2_{\mu}+\sigma^2_U+\sigma^2_{\omega}}}\right)\right) \\
  \Pr(T_i=d) & = (1-p_{\xi})\left(\Phi\left(\frac{\bar{y}-\bar{\mu}}{\sqrt{(1+\gamma^2)\sigma^2_{\mu}+\sigma^2_U+\sigma^2_{\omega}}}\right)
                  -\Phi\left(\frac{\bar{y}-\underline{\kappa}-\bar{\mu}}{\sqrt{(1+\gamma^2)\sigma^2_{\mu}+\sigma^2_U+\sigma^2_{\omega}}}\right)\right).
\end{align*}

Let's write functions to compute these objects:

```{r PropCompDef,eval=TRUE,echo=TRUE,results='hide'}
# proportion compliers
Prop.Comp <- function(param){
  first <- pnorm((log(param['barY'])+param['barkappa']-param['barmu'])/(sqrt((1+param['gamma']^2)*param['sigma2mu']+param['sigma2U']+param['sigma2omega'])))
  second <- pnorm((log(param['barY'])-param['barmu'])/(sqrt((1+param['gamma']^2)*param['sigma2mu']+param['sigma2U']+param['sigma2omega'])))
  return(param['pxi']*(first - second))  
}

# proportion defiers
Prop.Def <- function(param){
  first <- pnorm((log(param['barY'])-param['barmu'])/(sqrt((1+param['gamma']^2)*param['sigma2mu']+param['sigma2U']+param['sigma2omega'])))
  second <- pnorm((log(param['barY'])-param['underbarkappa']-param['barmu'])/(sqrt((1+param['gamma']^2)*param['sigma2mu']+param['sigma2U']+param['sigma2omega'])))
  return((1-param['pxi'])*(first - second))  
}

```

In our example, the proportion of compliers is equal to `r round(Prop.Comp(param),2)` and the proportion of defiers is equal to `r round(Prop.Def(param),2)`.
As a consequence, the population value of the numerator of the Wald estimator is equal to `r round(mean.alpha.compliers.pop[[1]]*Prop.Comp(param) - mean.alpha.defiers.pop[[1]]*Prop.Def(param),2)`.
In the Wald estimator, this quantity is divided by the difference between the proportion of participants when $Z_i=1$ and when $Z_i=0$. 
We have already computed this quantity earlier, but it is nice to try to compute it in a different way using the types.
The difference in the proportion of participants when $Z_i=1$ and when $Z_i=0$ is indeed equal to the difference in the proportion of compliers and the proportion of defiers. 
The difference between the proportion of compliers and the proportion of defiers is equal to `r round(Prop.Comp(param)-Prop.Def(param),2)`, while the difference between the proportion of participants when $Z_i=1$ and when $Z_i=0$ is equal to `r round(prob.D.Z.1(param)-prob.D.Z.0(param),2)`.
It is reassuring that we find the same thing (actually, full disclosure, I did not find the same thing at first, and this help me spot a mistake in the formulas for the proportions of participants: mistakes are normal and natural and that is how we learn and grow).

So we are now equipped to compute the value of the Wald estimator in the population in our model without monotonicity.
It is equal to `r round((mean.alpha.compliers.pop[[1]]*Prop.Comp(param) - mean.alpha.defiers.pop[[1]]*Prop.Def(param))/(prob.D.Z.1(param)-prob.D.Z.0(param)),3)`.
In practice, the bias of the Wald estimator is rather small for the average treatment effect on the compliers (remember that it is equal to `r round(mean.alpha.compliers.pop[[1]],3)`). 
In order to understand why, it is useful to see that the bias of the Wald estimator for the average treatment effect on the compliers is equal to:

\begin{align*}
  \esp{\Delta_i^Y|T_i=c}-\Delta^Y_{Wald} & = \esp{\Delta_i^Y|T_i=c} + (\esp{\Delta_i^Y|T_i=c}-\esp{\Delta_i^Y|T_i=d})\frac{\Pr(T_i=d)}{\Pr(T_i=c)-\Pr(T_i=d)},
\end{align*}

where the equality follows from Theorem \@ref(thm:ITELATE) and some algebra.
In the absence of Monotonicity, when the impact on defiers is smaller than the impact of compliers, the Wald estimator is baised upward for the effect on the compliers (as it happens in our example).
In a model in which the effect of the treatment is larger on defiers than on compliers, the Wald estimator is biased downwards for the effect on compliers because defiers make the outcome of the control group seem too good.
In the extreme, when $\esp{\Delta_i^Y|T_i=d}>\esp{\Delta_i^Y|T_i=c}(1+\frac{\Pr(T_i=c)-\Pr(T_i=d)}{\Pr(T_i=d)})$, the Wald estimator can be negative whereas the effects on compliers and on defiers are both positive.
This happens when the effect on defiers is $1+\frac{\Pr(T_i=c)-\Pr(T_i=d)}{\Pr(T_i=d)}$ times larger than the effect on compliers. 
In our case, that means that the effect on defiers should be `r round(1+(Prop.Comp(param)-Prop.Def(param))/Prop.Def(param),0)` times larger than the effect on compliers for the Wald estimator to be negative, that is to say the effect on defiers should be equal to `r round((1+(Prop.Comp(param)-Prop.Def(param))/Prop.Def(param))*mean.alpha.compliers.pop[[1]],2)`, really much much much larger than the effect on compliers.

From there, we are going to explore three strategies in order to identify some true effect of the treatment using the Wald estimator:

  * The first strategy has been recently proposed by [de Chaisemartin (2017)](https://drive.google.com/file/d/16XWlDECIvreM7l_NHe-JkXgyuPhFT1QG/view). 
  It is valid in a model without monotonicity.
  * The second strategy assumes that the heterogeneity in treatment effects is uncorrelated to the treatment.
  * The last strategy is due to Imbens and Angrist (1994) and assumes that Monotonicity holds.
  
Let's review these solutions in turn.

#### Identification without Monotonicity

The approach delineated by [de Chaisemartin (2017)](https://drive.google.com/file/d/16XWlDECIvreM7l_NHe-JkXgyuPhFT1QG/view) does not assume away non-monotonicity.
Clement instead assumes that we can divide the population of compliers in two-subpopulations: the **compliers-defiers** ($T_i=cd$) and the **surviving-compliers** ($T_i=sc$).
The main assumption in Clement's approach is that *(i)* the compliers-defiers are in the same proportion as the defiers and *(ii)* that the average effect of the treatment on the compliers defiers is equal as the average effect of the treatment on the defiers.
These two assumptions can be formalized as follows:

```{hypothesis,CD,name="Compliers-defiers"}
We assume that there exists as subpopulation of compliers that are in the same proportion as the defiers and for whom the average effect of the treatment is equal as the average effect of the treatment on the defiers: 
  
\begin{align*}
(T_i=c) & = (T_i=cd)\cup (T_i=sc) \\
\Pr(T_i=cd) & = \Pr(T_i=d) \\
\esp{Y^1_i-Y^0_i|T_i=cd} & = \esp{Y^1_i-Y^0_i|T_i=d}.
\end{align*}
```

The first equation in Assumption \@ref(hyp:CD) imposes that the compliers-defiers and the surviving-compliers are a partition of the population of compliers.
From Assumption \@ref(hyp:CD), we can prove the following theorem:

```{theorem,deChaise,name="Identification of the effect on the surviving-compliers"}
Under Assumptions \@ref(hyp:FirstStage), \@ref(hyp:ExclusionRestriction), \@ref(hyp:Independence) and \@ref(hyp:CD), the Wald estimator identifies the effect of the treatment on the surviving-compliers:
  
\begin{align*}
  \Delta^Y_{Wald} & = \Delta^Y_{sc},
\end{align*}
```

with:

\begin{align*}
  \Delta^Y_{Wald} & = \frac{\esp{Y_i|Z_i=1} - \esp{Y_i|Z_i=0}}{\Pr(D_i=1|Z_i=1)-\Pr(D_i=1|Z_i=0)}\\
  \Delta^Y_{sc} & = \esp{Y^1_i-Y^0_i|T_i=sc}.
\end{align*}

```{proof}
Under Assumptions \@ref(hyp:ExclusionRestriction) and \@ref(hyp:Independence), Theorems \@ref(thm:ITELATE) and \@ref(thm:ITEEncourag) imply that the numerator of the Wald estimator is equal to $\Delta^Y_{ITE}$ with:
  
\begin{align*}
  \Delta^Y_{ITE}  & = \esp{Y_i^{1}-Y_i^{0}|T_i=c}\Pr(T_i=c)-\esp{Y_i^{1}-Y_i^{0}|T_i=d}\Pr(T_i=d).
\end{align*}

Now, we have that the effect on compliers can be decomposed in the effect on surviving-compliers and the effect on compliers-defiers using the Law of Iterated Expectations and the fact that $T_i=sc \Rightarrow T_i=c$ and $T_i=cd \Rightarrow T_i=c$:
  
\begin{align*}
\Delta^Y_{c} & = \esp{Y_i^{1}-Y_i^{0}|T_i=sc}\Pr(T_i=sc|T_i=c)+\esp{Y_i^{1}-Y_i^{0}|T_i=cd}\Pr(T_i=cd|T_i=c),
\end{align*}

Now, using the fact that $\Pr(T_i=sc|T_i=c)\Pr(T_i=c)=\Pr(T_i=sc)$ and $\Pr(T_i=cd|T_i=c)\Pr(T_i=c)=\Pr(T_i=cd)$ (because $\Pr(A|B)\Pr(B)=\Pr(A\cap B)$ and $\Pr(A\cap B)=\Pr(A)$ if $A \Rightarrow B$), we have:
  
\begin{align*}
  \Delta^Y_{ITE} & = \esp{Y_i^{1}-Y_i^{0}|T_i=sc}\Pr(T_i=sc)\\
                & \phantom{=}+\esp{Y_i^{1}-Y_i^{0}|T_i=cd}\Pr(T_i=cd)-\esp{Y_i^{1}-Y_i^{0}|T_i=d}\Pr(T_i=d).
\end{align*}

The second part of the right-hand side of the above equation is equal to zero by virtue of Assumption \@ref(hyp:CD).
Now, under Assumptions \@ref(hyp:FirstStage), \@ref(hyp:ExclusionRestriction) and \@ref(hyp:Independence), we know, from the proof of Theorem \@ref(thm:IdentLATE), that $\Pr(D_i=1|Z_i=1)-\Pr(D_i=1|Z_i=0)=\Pr(T_i=c)-\Pr(T_i=d)$.
Under Assumption \@ref(hyp:CD), we have $\Pr(T_i=c)=\Pr((T_i=cd)\cup(T_i=sc))=\Pr(T_i=cd)+\Pr(T_i=sc)$.
Replacing $\Pr(T_i=c)$ gives $\Pr(D_i=1|Z_i=1)-\Pr(D_i=1|Z_i=0)=\Pr(T_i=sc)$.
Dividing $\Delta^Y_{ITE}$ by $\Pr(T_i=sc)$ gives the result.
```

```{remark}
[de Chaisemartin (2017)](https://drive.google.com/file/d/16XWlDECIvreM7l_NHe-JkXgyuPhFT1QG/view) shows in his Theorem 2.1 that the reciprocal of Theorem \@ref(thm:deChaise) is actually valid: if there exists surviving-compliers such that their effect is estimated by the Wald estimator and their proportion is equal to the denominator of the Wald estimator, then it has to be that there exists a sub-population of compliers-defiers that are in the same proportion as the defiers and have the same average treatment effect.
```
  
```{example}  
Let us now see if the conditions in [de Chaisemartin (2017)](https://drive.google.com/file/d/16XWlDECIvreM7l_NHe-JkXgyuPhFT1QG/view) are verified in our numerical example.
```
I have bad news: they are not. 
It is not super easy to see why, but an intuitive explanation is that the average effect on the defiers in our model is taken conditional on $y^B_i+V_i\in]\bar{y}-\underline{\kappa},\bar{y}]$ while the effect on compliers is taken conditional on $y^B_i+V_i\in]\bar{y},\bar{y}+\bar{\kappa}]$. 
These two intervals do not overlap. 
Since the expected value of the treatment effect conditional on $y^B_i+V_i=v$ is monotonous in $v$ (because both variables come from a bivariate normal distribution), then all the effects on the defiers interval are either smaller or larger than all the effects on the compliers interval, making it impossible to find a sub-population of compliers that have the same average effect of the treatment as the defiers.

More formally, it is possible to prove this result by using the concept of Marginal Treatment Effect developed by [Heckman and Vytlacil (1999)](https://www.pnas.org/content/pnas/96/8/4730.full.pdf). 
I might devote a specific section of the book to the MTE and its derivations. 
For now, I let it as a possibility.

What can we do then?
Probably the best that we can do is to find $\kappa^*$ such that $\Pr(\bar{y}<y_i^B+V\leq\bar{y}+\kappa^*)p_{\xi}=\Pr(T_i=d)$, that is the value such that the interval of values of $y_i^B+V$ that are for compliers and closest to the interval for defiers and that contains the same proportion of compliers as there are defiers.
This value is going to produce an average effect for compliers-defiers as close as possible to the average effect on defiers. 
It can be computed as follows:

\begin{align*}
  \kappa^* & = \bar{\mu}-\bar{y}+\sqrt{(1+\gamma^2)\sigma^2_{\mu}+\sigma^2_U+\sigma^2_{\omega}}
                                  \Phi^{-1}\Bigg(\Phi\left(\frac{\bar{y}-\bar{\mu}}{\sqrt{(1+\gamma^2)\sigma^2_{\mu}+\sigma^2_U+\sigma^2_{\omega}}}\right)\\
           & \phantom{=}+\frac{1-p_{\xi}}{p_{\xi}}\left(\Phi\left(\frac{\bar{y}-\bar{\mu}}{\sqrt{(1+\gamma^2)\sigma^2_{\mu}+\sigma^2_U+\sigma^2_{\omega}}}\right)-\Phi\left(\frac{\bar{y}-\underline{\kappa}-\bar{\mu}}{\sqrt{(1+\gamma^2)\sigma^2_{\mu}+\sigma^2_U+\sigma^2_{\omega}}}\right)\right)\Bigg)
 \end{align*}

Let's write functions to compute $\kappa^*$, the implied proportion of compliers-defiers and the average effect of the treatment on compliers-defiers and on surviving-compliers:

```{r KappaStar,eval=TRUE,echo=TRUE,results='hide'}
# kappa star
KappaStar <- function(param){
  prop.def <- Prop.Def(param)
  prop.below.bary <- pnorm((log(param['barY'])-param['barmu'])/(sqrt((1+param['gamma']^2)*param['sigma2mu']+param['sigma2U']+param['sigma2omega'])))
  st.dev.yB.V <- sqrt((1+param['gamma']^2)*param['sigma2mu']+param['sigma2U']+param['sigma2omega'])
  return(param['barmu']-log(param['barY'])+st.dev.yB.V*qnorm(prop.below.bary+prop.def/param['pxi']))
}

# proportion of compliers-defiers
Prop.Comp.Def <- function(param){
  first <- pnorm((log(param['barY'])+KappaStar(param)-param['barmu'])/(sqrt((1+param['gamma']^2)*param['sigma2mu']+param['sigma2U']+param['sigma2omega'])))
  second <- pnorm((log(param['barY'])-param['barmu'])/(sqrt((1+param['gamma']^2)*param['sigma2mu']+param['sigma2U']+param['sigma2omega'])))
  return(param['pxi']*(first - second))  
}

# mean impact on compliers-defiers
lower.cut.comp.def <- c(-Inf,log(param['barY']))
upper.cut.comp.def <- c(Inf,log(param['barY'])+KappaStar(param))
mean.alpha.comp.def.pop <- mtmvnorm(mean=mean.alpha.yBV,sigma=cov.alpha.yBV,lower=lower.cut.comp.def,upper=upper.cut.comp.def,doComputeVariance=FALSE)[[1]]

# mean impact on surviving compliers
lower.cut.surv.comp <- c(-Inf,log(param['barY'])+KappaStar(param))
upper.cut.surv.comp <- c(Inf,log(param['barY'])+param['barkappa'])

mean.alpha.surv.comp.pop <- mtmvnorm(mean=mean.alpha.yBV,sigma=cov.alpha.yBV,lower=lower.cut.surv.comp,upper=upper.cut.surv.comp,doComputeVariance=FALSE)[[1]]

```

The first have that $\kappa^*=$ `r round(KappaStar(param),4)`. 
For this value of $\kappa^*$, we have that $\Pr(T_i=cd)=$ `r round(Prop.Comp.Def(param),4)`.
As expected, this is very close to the proportion of compliers in the population: $\Pr(T_i=d)=$ `r round(Prop.Def(param),4)`.
Finally, the average treatment effect on the compliers-defiers is equal to: $\Delta^y_{cd}=$ `r round(mean.alpha.comp.def.pop[[1]],4)`.
As expected, but luckily enough, since it was absolutely not sure, it is very close to the to the average treatment effect on the defiers: $\Delta^y_{d}=$ `r round(mean.alpha.defiers.pop[[1]],4)`.
So, in our model, Assumption \@ref(hyp:CD) is almost satisfied, and so does Theorem \@ref(thm:deChaise).
As a consequence, the Wald estimator is very close to the effect on the surviving-compliers.
Indeed, the Wald estimator, in the population, is equal to $\Delta^y_{Wald}=$ `r round((mean.alpha.compliers.pop[[1]]*Prop.Comp(param) - mean.alpha.defiers.pop[[1]]*Prop.Def(param))/(prob.D.Z.1(param)-prob.D.Z.0(param)),6)`, while the average effect on surviving-compliers is equal to $\Delta^y_{sc}=$ `r round(mean.alpha.surv.comp.pop[[1]],6)`.

#### Identification under Independence of treatment effects

Another way to get around the issue of Non-Monotonicity is simply to assume away any meaningful role for treatment effect heterogeneity.
One approach to that would simply be to assume that treatment effects are constant across individuals.
I leave to the reader to prove that in that case, the Wald estimator would recover the treatment effect under only Independence and Exclusion Restriction.
We are going to use a slightly more general approach here by assuming that treatment effect heterogeneity is unrelated to the reaction to the instrument:

```{hypothesis,IndepTreatEffect,name="Independent Treatment Effects"}
We assume that the treatment effect is independent from potential reactions to the instrument:

\begin{align*}
\Delta^Y_i\Ind (D^1_i,D^0_i).
\end{align*}
```

We can now prove that, under Assumption \@ref(hyp:IndepTreatEffect), the Wald estimator identifies the Average Treatment Effect (ATE), the average effect of the Treatment on the Treated (TT) and the average effect on compliers and on defiers.
The first thing to know before we state the result is that, under Assumption \@ref(hyp:IndepTreatEffect), all these average treatment effects are equal to each other.
This is a direct implication of the following lemma:

```{lemma,IndepTreatEffectType,name="Independence of Treatment Effects from Types"}
Under Assumption \@ref(hyp:IndepTreatEffect), the treatment effect is independent from types:
  
\begin{align*}
\Delta^Y_i\Ind T_i.
\end{align*}
```

```{proof}
Lemma (4.2) in [Dawid (1979)](https://rss.onlinelibrary.wiley.com/doi/10.1111/j.2517-6161.1979.tb01052.x) states that if $X \Ind Y|Z$ and $U$ is a function of $X$, then $U \Ind Y|Z$.
Since $T_i$ is a function of $(D^1_i,D^0_i)$ under Assumption \@ref(hyp:IndepTreatEffect), Lemma \@ref(lem:IndepTreatEffectType) follows. 
```

A direct corollary of Lemma \@ref(lem:IndepTreatEffectType) is:

```{corollary,IndepTreatEffectAve,name="Independence of Treatment Effects and Average Effects"}
Under Assumption \@ref(hyp:IndepTreatEffect), the Average Treatment Effect (ATE), the average effect of the Treatment on the Treated (TT) and the average effect on compliers and on defiers are all equal:
  
\begin{align*}
\Delta^Y_{ATE} = \Delta^Y_{TT(1)} = \Delta^Y_{TT(0)} = \Delta^Y_{c} = \Delta^Y_{d}.
\end{align*}
```

with:
\begin{align*}
\Delta^Y_{TT(z)} = \esp{Y_i^1-Y_i^0|D_i=1,Z_i=z}.
\end{align*}

```{proof}
Using Lemma \@ref(lem:IndepTreatEffectType), we have that:

\begin{align*}
  \Delta^Y_{c} = \Delta^Y_{d} = \Delta^Y_{at} =\Delta^Y_{nt}.
\end{align*}

Because $T_i$ is a partition, we have $\Delta^Y_{ATE}=\Delta^Y_{c}\Pr(T_i=c)+\Delta^Y_{d}\Pr(T_i=d)+\Delta^Y_{at}\Pr(T_i=at)+\Delta^Y_{nt}\Pr(T_i=nt)=\Delta^Y_{c}$ (since $\Pr(T_i=c)+\Pr(T_i=d)+\Pr(T_i=at)+\Pr(T_i=nt)=1$).
Finally, we also have that $\Delta^Y_{TT(1)}=\Delta^Y_{c}\Pr(T_i=c|D_i=1,Z_i=1)+\Delta^Y_{at}\Pr(T_i=at|D_i=1,Z_i=1)=\Delta^Y_{c}$ and $\Delta^Y_{TT(0)}=\Delta^Y_{d}\Pr(T_i=d|D_i=1,Z_i=0)+\Delta^Y_{at}\Pr(T_i=at|D_i=1,Z_i=0)=\Delta^Y_{c}$, since $(D_i=1)\cap(Z_i=1)\Rightarrow (T_i=c)\cup(T_i=at)$ and $(D_i=1)\cap(Z_i=0)\Rightarrow (T_i=d)\cup(T_i=at)$.
```

We are now equioped to state the final result of this section:

```{theorem,IdentIndepTreatEffect,name="Identification under Independent Treatment Effect"}
Under Assumptions \@ref(hyp:FirstStage), \@ref(hyp:ExclusionRestriction), \@ref(hyp:Independence) and \@ref(hyp:IndepTreatEffect), the Wald estimator identifies the average effect of the Treatment on the Treated:
  
\begin{align*}
  \Delta^Y_{Wald} & = \Delta^Y_{TT}.
\end{align*}
```

```{proof}
Using the formula for the Wald estimator, we have, for the two components of its numerator:

\begin{align*}
\esp{Y_i|Z_i=1} & = \esp{Y_i^0+(Y_i^1-Y_i^0)D_i|Z_i=1} \\
                & = \esp{Y_i^0|Z_i=1}+\esp{\Delta^Y_i|D_i=1,Z_i=1}\Pr(D_i=1|Z_i=1)\\
                & = \esp{Y_i^0|Z_i=1}+\Delta^Y_{TT(1)}\Pr(D_i=1|Z_i=1)\\
\esp{Y_i|Z_i=0} & = \esp{Y_i^0+(Y_i^1-Y_i^0)D_i|Z_i=0} \\
                & = \esp{Y_i^0|Z_i=0}+\esp{\Delta^Y_i|D_i=0,Z_i=1}\Pr(D_i=1|Z_i=0)\\
                & = \esp{Y_i^0|Z_i=0}+\Delta^Y_{TT(0)}\Pr(D_i=1|Z_i=0),\\
\end{align*}

where the first equalities use Assumption \@ref(hyp:ExclusionRestriction).
Now, under Assumption \@ref(hyp:IndepTreatEffect), Corollary \@ref(cor:IndepTreatEffectAve) implies that $\Delta^Y_{TT(0)}=\Delta^Y_{TT(1)}=\Delta^Y_{TT}$.
We thus have that the numerator of the Wald estimator is equal to:

\begin{align*}
\esp{Y_i|Z_i=1}-\esp{Y_i|Z_i=0} & = \Delta^Y_{TT}(\Pr(D_i=1|Z_i=1)-\Pr(D_i=1|Z_i=0))\\
                                & \phantom{=}+\esp{Y_i^0|Z_i=1}-\esp{Y_i^0|Z_i=0}.
\end{align*}

Assumption \@ref(hyp:Independence) implies that $\esp{Y_i^0|Z_i=1}=\esp{Y_i^0|Z_i=0}$.
Using Assumption \@ref(hyp:FirstStage) proves the result.
```

#### Identification under Monotonicity

The classical approach to identification using instrumental variables is due to [Imbens and Angrist (1994)](https://www.jstor.org/stable/2951620) and [Angrist, Imbens and Rubin (1996)](https://www.jstor.org/stable/2291629). 
It rests on Assumption \@ref(hyp:MonotonicityIV) or Monotonicity that we are now familiar with, that requires that the effect of the instrument on treatment participation moves everyone in the same direction.

```{remark}
For the rest of the section, we will assume  that $\forall i$, $D^1_i\geq D_i^0$.
It is without loss of generality, since if the initial treatment does not comply with this requirement, you can simply redefine a new treatment equal to $-D_i$.
```

Under Monotonicity, there are no defiers.
This is what the following lemma shows:

```{lemma,NoDefiers,Name="No Defiers"}
Under Assumption \@ref(hyp:MonotonicityIV), there are no defiers a.s.: 
  
\begin{align*}
  \Pr(T_i=c) & = 0.
\end{align*}
```

```{proof}
Under Assumption \@ref(hyp:MonotonicityIV), $\forall i$, $D^1_i\geq D_i^0$.
As a consequence, $\Pr(D^1_i < D_i^0)=0$.
Since defiers are defined as $D^1_i < D_i^0$, the result follows.
```

In the absence of defiers, the Wald estimator identifies the average effect of the treatment on the compliers, also called the Local Average Treatment Effect:

```{theorem,IdentLATEIV,Name="Identification of the Local Average Treatment Effect with Instrumental Variables"}
Under Assumptions  \@ref(hyp:FirstStage), \@ref(hyp:ExclusionRestriction), \@ref(hyp:Independence) and \@ref(hyp:MonotonicityIV), the Wald estimator identifies the average effect of the treatment on the compliers, also called the Local Average Treatment Effect: 
  
\begin{align*}
  \Delta^Y_{Wald}& = \Delta^Y_{LATE}.
\end{align*}
```

```{proof}
Using Theorem \@ref(thm:IdentLATE) directly proves the result.
```

```{remark}
The magic of the instrumental variables setting applies again.
By moving the instrument, we are able to learn something about the causal effect of the treatment. 
Monotonicity is a very strong assumption though, as are Independence and Exclusion Restriction. 
They are very rarely met in practice.
Even the case of RCTs with encouragement design, where Independence holds by design, might be affected by failures of Exclusion Restriction and/or Monotonicity.
```

```{example}
Let's see how monotonicity works in our example.
```

First, we have to generate a model in which monotonicity holds.
For that, we need to shut down heterogeneous reactions to the instrument.
In practice, we are going to replace the participation equation in our model, which was characterized by a random coefficient, by the following one, which has a constant coefficient:

\begin{align*}
  D_i & = \uns{y_i^B-\bar{\kappa} Z_i + V_i\leq\bar{y}} 
\end{align*}

As a consequence, we have no more defiers and monotonicity holds.
Let us now generate the data from the model with monotonicity:

```{r simulIVMono,eval=TRUE,echo=TRUE,results='hide'}
set.seed(12345)
N <-1000
cov.eta.omega <- matrix(c(param["sigma2eta"],param["rhoetaomega"]*sqrt(param["sigma2eta"]*param["sigma2omega"]),param["rhoetaomega"]*sqrt(param["sigma2eta"]*param["sigma2omega"]),param["sigma2omega"]),ncol=2,nrow=2)
eta.omega <- as.data.frame(mvrnorm(N,c(0,0),cov.eta.omega))
colnames(eta.omega) <- c('eta','omega')
mu <- rnorm(N,param["barmu"],sqrt(param["sigma2mu"]))
UB <- rnorm(N,0,sqrt(param["sigma2U"]))
yB <- mu + UB 
YB <- exp(yB)
Ds <- rep(0,N)
Z <- rbinom(N,1,param["pZ"])
V <- param["gamma"]*(mu-param["barmu"])+eta.omega$omega
Ds[yB-param["barkappa"]*Z+V<=log(param["barY"])] <- 1 
epsilon <- rnorm(N,0,sqrt(param["sigma2epsilon"]))
U0 <- param["rho"]*UB + epsilon
y0 <- mu +  U0 + param["delta"]
alpha <- param["baralpha"]+  param["theta"]*mu + eta.omega$eta
y1 <- y0+alpha
Y0 <- exp(y0)
Y1 <- exp(y1)
y <- y1*Ds+y0*(1-Ds)
Y <- Y1*Ds+Y0*(1-Ds)
```

We can now define the types variable $T_i$:

```{r typesillustrationMono,eval=TRUE,echo=TRUE,results='hide'}
D1 <- ifelse(yB-param["barkappa"]+V<=log(param["barY"]),1,0)
D0 <- ifelse(yB+V<=log(param["barY"]),1,0)
AT <- ifelse(D1==1 & D0==1,1,0)
NT <- ifelse(D1==0 & D0==0,1,0)
C <- ifelse(D1==1 & D0==0,1,0)
D <- ifelse(D1==0 & D0==1,1,0)
Type <- ifelse(AT==1,'a',
            ifelse(NT==1,'n',
                   ifelse(C==1,'c',
                          ifelse(D==1,'d',""))))

data.mono <- data.frame(cbind(Type,C,NT,AT,D1,D0,Y,y,Y1,Y0,y0,y1,yB,alpha,U0,eta.omega$eta,epsilon,Ds,Z,mu,UB))
```

The first thing we can check is that there are no defiers. 
For that, let's count the number of individuals who have $T_i=1$.
It is equal to `r sum(Type=="d")`.

One thing that helped me understand how the IV approach under monotonicity works is the following graph:

```{r plottypesMono,eval=TRUE,echo=TRUE,fig.cap='Types under Monotonicity',fig.align='center',out.width='50%',fig.pos='htbp'}
plot(yB[AT==1]+V[AT==1],y[AT==1],pch=1,xlim=c(5,11),ylim=c(5,11),xlab="yB+V",ylab="Outcomes")
points(yB[NT==1]+V[NT==1],y[NT==1],pch=1,col='blue')
points(yB[C==1 & Ds==1]+V[C==1 & Ds==1],y[C==1 & Ds==1],pch=1,col='red')
points(yB[C==1 & Ds==0]+V[C==1 & Ds==0],y[C==1 & Ds==0],pch=1,col='green')
abline(v=log(param["barY"]),col="red")
abline(v=log(param["barY"])+param['barkappa'],col="red")
text(x=c(log(param["barY"]),log(param["barY"])+param['barkappa']),y=c(5,5),labels=c(expression(bar('y')),expression(bar('y')+bar(kappa))),pos=c(2,4),col=c("red","red"))
legend(5,10.5,c('AT','NT','C|D=1','C|D=0'),pch=c(1,1,1,1),col=c("black",'blue',"red",'green'),ncol=1)
```

What \@ref(fig:plottypesMono) shows is that the IV acts as a randomized controlled trial among compliers.
Within the population of compliers, whether one receives the treatment or not is as good as random.
If we actually knew who the compliers were, we could directly estimate the effect of the treatment by comparing the outcomes of the treated compliers to the outcomes of the untreated compliers. 
Actually, this approach, applied in our sample, yields an estimated treatment effect on the compliers of `r round(mean(y[C==1 & Ds==1])-mean(y[C==1 & Ds==0]),2)`, whereas the simple comparison of participants and non participants would give an estimate of  `r round(mean(y[Ds==1])-mean(y[Ds==0]),2)`.
In our sample, the average effect of the treatment on compliers is actually equal to  `r round(mean(alpha[C==1]),2)`.

Let us finally check that Theorem \@ref(thm:IdentLATEIV) works in the population in our new model.
We need to compute the various parts of the Wald estimator and the average effect of the treatment on the compliers.
The key to understand the Wald estimator is to see that its numerator is composed of the difference between two means, with both means containing the average outcomes of always takers and never takers weighted by their respective proportions in the population, as shown in the proof of Theorem \@ref(thm:IdentLATEIV).
These two means cancel out, leaving only the differences in the means of the compliers in and out of the treatment, weighted by their proportion in the population.
The denominator of the Wald estimator simply provides an estimate of the proportion of compliers. 
In order to illustrate these intuitions in our example, I am going to use the formula for a truncated multivariate normal variable and the package `tmvtnorm`.
The most important thing to notice here is that $(y^0_i,y^1_i,y_i^B+V_i) \sim \mathcal{N}\left(\bar{\mu}+\delta,\bar{\mu}(1+\theta)+\delta+\bar{\alpha},\bar{\mu},\mathbf{C}\right)$ with:

\begin{align*}
  \mathbf{C} &=                    \left(\begin{array}{ccc}
                                              \sigma^2_{\mu}+\rho^2\sigma^2_{U} +\sigma^2_{\epsilon} &
                                              (1+\theta)\sigma^2_{\mu}+\rho^2\sigma^2_U + \sigma^2_{\epsilon} &
                                              (1+\gamma)\sigma^2_{\mu}+\rho\sigma^2_U \\
                                              (1+\theta)\sigma^2_{\mu}+\rho^2\sigma^2_U + \sigma^2_{\epsilon} &
                                              (1+\theta^2)\sigma^2_{\mu}+\rho^2\sigma^2_{U} +\sigma^2_{\epsilon} + \sigma^2_{\eta} &
                                              (1+\theta+\gamma)\sigma^2_{\mu}+\rho\sigma^2_U+\rho_{\eta,\omega}\sigma^2_{\eta}\sigma^2_{\omega} \\
                                              (1+\gamma)\sigma^2_{\mu}+\rho\sigma^2_U &
                                              (1+\theta+\gamma)\sigma^2_{\mu}+\rho\sigma^2_U+\rho_{\eta,\omega}\sigma^2_{\eta}\sigma^2_{\omega} &
                                              (1+\gamma^2)\sigma^2_{\mu}+\sigma^2_U+\sigma^2_{\omega} \\
                                              \end{array}
                                        \right)
\end{align*}

We now simply have to derive the mean outcomes and proportions of each type in the population in order to form the Wald estimator.
Let me first derive the joint distribution of the portential outcomes and the means and proportions of each type in the population.

```{r OutcomesTypesPopulation,eval=TRUE,echo=TRUE,results='hide'}
mean.y0.y1.yBV <- c(param['barmu']+param['delta'],param['barmu']*(1+param['theta'])+param['delta']+param['baralpha'],param['barmu'])
cov.y0.y1.yBV <- matrix(c(param['sigma2mu']+param['rho']^2*param['sigma2U']+param['sigma2epsilon'],
                          (1+param['theta'])*param['sigma2mu']+param['rho']^2*param['sigma2U']+param['sigma2epsilon'],
                          (1+param['gamma'])*param['sigma2mu']+param['rho']*param['sigma2U'],
                          (1+param['theta'])*param['sigma2mu']+param['rho']^2*param['sigma2U']+param['sigma2epsilon'],
                          (1+param['theta']^2)*param['sigma2mu']+param['rho']^2*param['sigma2U']+param['sigma2epsilon']+param['sigma2eta'],
                          (1+param['theta']+param['gamma'])*param['sigma2mu']+param['rho']*param['sigma2U']+param['rhoetaomega']*param['sigma2eta']*param['sigma2omega'],
                          (1+param['gamma'])*param['sigma2mu']+param['rho']*param['sigma2U'],
                          (1+param['theta']+param['gamma'])*param['sigma2mu']+param['rho']*param['sigma2U']+param['rhoetaomega']*param['sigma2eta']*param['sigma2omega'],
                          (1+param['gamma']^2)*param['sigma2mu']+param['sigma2U']+param['sigma2omega']),3,3,byrow=TRUE)

# cuts
#always takers
lower.cut.at <- c(-Inf,-Inf,-Inf)
upper.cut.at <- c(Inf,Inf,log(param['barY']))
# compliers
lower.cut.comp <- c(-Inf,-Inf,log(param['barY']))
upper.cut.comp <- c(Inf,Inf,log(param['barY'])+param['barkappa'])
# never takers
lower.cut.nt <- c(-Inf,-Inf,log(param['barY'])+param['barkappa'])
upper.cut.nt <- c(Inf,Inf,Inf)

# means by types
#always takers
mean.y1.at.pop <- mtmvnorm(mean=mean.y0.y1.yBV,sigma=cov.y0.y1.yBV,lower=lower.cut.at,upper=upper.cut.at,doComputeVariance=FALSE)[[1]][[2]]
mean.y0.at.pop <- mtmvnorm(mean=mean.y0.y1.yBV,sigma=cov.y0.y1.yBV,lower=lower.cut.at,upper=upper.cut.at,doComputeVariance=FALSE)[[1]][[1]]
# never takers
mean.y1.nt.pop <- mtmvnorm(mean=mean.y0.y1.yBV,sigma=cov.y0.y1.yBV,lower=lower.cut.nt,upper=upper.cut.nt,doComputeVariance=FALSE)[[1]][[2]]
mean.y0.nt.pop <- mtmvnorm(mean=mean.y0.y1.yBV,sigma=cov.y0.y1.yBV,lower=lower.cut.nt,upper=upper.cut.nt,doComputeVariance=FALSE)[[1]][[1]]
#compliers
mean.y1.comp.pop <- mtmvnorm(mean=mean.y0.y1.yBV,sigma=cov.y0.y1.yBV,lower=lower.cut.comp,upper=upper.cut.comp,doComputeVariance=FALSE)[[1]][[2]]
mean.y0.comp.pop <- mtmvnorm(mean=mean.y0.y1.yBV,sigma=cov.y0.y1.yBV,lower=lower.cut.comp,upper=upper.cut.comp,doComputeVariance=FALSE)[[1]][[1]]

# Proportion of each types
# always takers
prop.at.pop <- ptmvnorm.marginal(log(param['barY']),n=3,mean=mean.y0.y1.yBV,sigma=cov.y0.y1.yBV)[[1]]
# never takers
prop.nt.pop <- 1-ptmvnorm.marginal(log(param['barY'])+param['barkappa'],n=3,mean=mean.y0.y1.yBV,sigma=cov.y0.y1.yBV)[[1]]
# compliers
prop.comp.pop <- ptmvnorm.marginal(log(param['barY'])+param['barkappa'],n=3,mean=mean.y0.y1.yBV,sigma=cov.y0.y1.yBV)[[1]]-ptmvnorm.marginal(log(param['barY']),n=3,mean=mean.y0.y1.yBV,sigma=cov.y0.y1.yBV)[[1]]

# LATE
late.pop <- mean.y1.comp.pop-mean.y0.comp.pop
late.prop.comp.pop <- late.pop*prop.comp.pop
# Wald
num.Wald.pop <- (mean.y1.comp.pop*prop.comp.pop+mean.y1.at.pop*prop.at.pop+mean.y0.nt.pop*prop.nt.pop-(mean.y0.comp.pop*prop.comp.pop+mean.y1.at.pop*prop.at.pop+mean.y0.nt.pop*prop.nt.pop))
denom.Wald.pop <- (prop.at.pop+prop.comp.pop-prop.at.pop)
Wald.pop <- num.Wald.pop/denom.Wald.pop
```

We are now equipped to compute the Wald estimator in the population.
Before that, let us compute the LATE.
We have $\Delta^Y_{LATE} =$ `r round(late.pop,3)`.
The Wald estimator is equal to $\Delta^Y_{Wald} =$ `r round(Wald.pop,3)`.
They are obviously equal.
This is because the numerator of the Wald is equal to the product of the LATE multiplied by the proportion of compliers (which is equal to `r round(late.prop.comp.pop,3)`).
This is because the outcomes of never takers and always takers cancel out on each separate term of the numerator of the Wald estimator.
Indeed, we have that the numerator of the Wald estimator is equal to: `r round(num.Wald.pop,3)`.

### Estimation

Estimation of the LATE under the IV assumptions closely follows the same steps that we have delineated in Section \@ref(IVRCT):

  1. **First stage** regression of $D_i$ on $Z_i$: this estimates the impact of the instrument on participation into the program and estimates the proportion of compliers.
  2. **Reduced form** regression of $Y_i$ on $Z_i$: this estimates the impact of the instrument on outcomes, *a.k.a* the ITE.
  3. **Structural** regression of $Y_i$ on $D_i$ using $Z_i$ as an instrument, which estimates the LATE.

Let's take these three steps in turn.

#### First stage regression

The first stage regression regresses $D_i$ on $Z_i$ and thus estimates the impact of the instrument on treatment participation, which is equal to the proportion of compliers.
It can be run using the With/Without estimator or OLS (both are numerically equivalent as Lemma \@ref(lem:WWOLS) shows) or OLS conditioning on observed covariates.

```{example}
Let's see how these three approaches fare in our example.
```

```{r FirstStageIV,eval=TRUE,echo=TRUE,results='hide'}
# WW first stage
WW.First.Stage.IV <- mean(Ds[Z==1])-mean(Ds[Z==0])
# Simple OLS
OLS.D.Z.IV <- lm(Ds~Z)
OLS.First.Stage.IV <- coef(OLS.D.Z.IV)[[2]]
# OLS conditioning on yB
OLS.D.Z.yB.IV <- lm(Ds~Z+yB)
OLSX.First.Stage.IV <- coef(OLS.D.Z.yB.IV)[[2]]
```

The WW estimator of the first stage impact of $Z_i$ on $D_i$ is equal to `r round(WW.First.Stage.IV,3)`.
The OLS estimator of the first stage impact of $Z_i$ on $D_i$ is equal to `r round(OLS.First.Stage.IV,3)`.
The OLS estimator of the first stage impact of $Z_i$ on $D_i$ conditioning on $y^B_i$ is equal to `r round(OLSX.First.Stage.IV,3)`.
Remember that the true proportion of compliers in the population in our model is equal to `r round(prop.comp.pop,3)`.

#### Reduced form regression

The reduced form regression regresses $Y_i$ on $Z_i$ and thus estimates the impact of the instrument on outcomes, which is equal to the ITE.
It can be run using the With/Without estimator or OLS (both are numerically equivalent as Lemma \@ref(lem:WWOLS) shows) or OLS conditioning on observed covariates.

```{example}
Let's see how these three approaches fare in our example.
```

```{r ReducedFormIV,eval=TRUE,echo=TRUE,results='hide'}
# WW reduced form
WW.Reduced.Form.IV <- mean(y[Z==1])-mean(y[Z==0])
# Simple OLS
OLS.y.Z.IV <- lm(y~Z)
OLS.Reduced.Form.IV <- coef(OLS.y.Z.IV)[[2]]
# OLS conditioning on yB
OLS.y.Z.yB.IV <- lm(y~Z+yB)
OLSX.Reduced.Form.IV <- coef(OLS.y.Z.yB.IV)[[2]]
```

The WW estimator of the reduced form impact of $Z_i$ on $y_i$ is equal to `r round(WW.Reduced.Form.IV,3)`.
The OLS estimator of the reduced form impact of $Z_i$ on $y_i$ is equal to `r round(OLS.Reduced.Form.IV,3)`.
The OLS estimator of the reduced form impact of $Z_i$ on $y_i$ conditioning on $y^B_i$ is equal to `r round(OLSX.Reduced.Form.IV,3)`.
Remember that the true ITE in the population in our model is equal to `r round(late.prop.comp.pop,3)`.

#### Structural regression

The final step of the analysis is to estimate the impact of $D_i$ on $Y_i$ using $Z_i$ as an instrument.
This can be done either by directly using the Wald estimator, by dividing the estimate of the reduced form by the result of the first stage, or by directly using the IV estimator (which is equivalent to the Wald estimator as Theorem \@ref(thm:WaldIV) shows) or the IV estimator conditional on covariates.

```{example}
Let's see how these four approaches fare in our example.
```

```{r StructuralFormIV,eval=TRUE,echo=TRUE,results='hide'}
# Wald structural form
Wald.Structural.Form.IV <- (mean(y[Z==1])-mean(y[Z==0]))/(mean(Ds[Z==1])-mean(Ds[Z==0]))
# Simple IV
TSLS.y.D.Z.IV <- ivreg(y~Ds|Z)
TSLS.Structural.Form.IV <- coef(TSLS.y.D.Z.IV)[[2]]
# IV conditioning on yB
TSLS.y.D.Z.yB.IV <- ivreg(y~Ds+yB|Z+yB)
TSLSX.Structural.Form.IV <- coef(TSLS.y.D.Z.yB.IV)[[2]]
```

The Wald estimator of the LATE is equal to $\hat{\Delta}_{Wald}^{y}=$ `r round(Wald.Structural.Form.IV,3)`.
The IV estimator of the LATE is equal to $\hat{\Delta}_{IV}^{y}=$ `r round(TSLS.Structural.Form.IV,3)`, and is numerically identical to the Wald estimator, as expected.
The IV estimator of the LATE conditioning on $y_i^B$ is equal to `r round(TSLSX.Structural.Form.IV,3)`.
Remember that the true LATE in the population in our model is equal to `r round(late.pop,3)`.

```{remark}
The last thing we might want to check is what the sampling noise of the IV estimator looks like and whether it is reduced by conditioning on observed covariates.
```

```{example}
Let's see how sampling noise moves in our example.
```

**Do it**


### Estimation of sampling noise



```{remark}
The framework we have seen here as been extended to multivalued instruments or treatments by several papers.
[Imbens and Angrist (1994)](https://www.jstor.org/stable/2951620) extend the framework to an ordered instrument.
They show that the 2SLS estimator is a weighted average of LATEs for each values of the instrument, with positive weights summing to one.
[Angrist and Imbens (1995)](https://www.tandfonline.com/doi/abs/10.1080/01621459.1995.10476535) extend the framework to he case where the treatment is an ordered discrete variable and there are multiple dichotomous instruments.
They again show that the 2SLS estimator is a weighted average of LATEs with positive weights summing to one. 
[Heckman and Vytlacil (1999)](https://www.pnas.org/content/96/8/4730) extend the framework to a case with a continuous instrument and show that one can the define a Marginal Treatment Effect (or MTE) that is equal to the effect of the treatment on individuals that have the same unobserved propensity to take the treatment.
They show that the MTE can be identified by a limiting form of Wald estimator that they call a Local Instrumental Variable estimator.
They also show that average treatment effect parameters such as TT, ATE and LATE are all weighted averages of the MTE, with positive weights summing to one. 
Under strong support conditions on the side of the instrument, one can thus in principle recover all treatment effect parameters with a continuous instrument.
```

```{remark}
One important concern with the first stage regression is that of weak instruments. 
When Assumption \@ref(hyp:FirstStage) does not hold and the impact on the instrument on take up is actually zero in the population, the Wald estimator is not well-defined.
```

**Expand**

## Regression Discontinuity Designs



## Difference In Differences

In Difference In Differences (a.k.a. DID), the difference between treated and untreated before the treatment is used to approximate selection bias.
As a consequence, DID works by correcting the With/Without comparison after treatment by the With/Without comparison before treatment and hopes that it is enough to recover the TT.
Hence the name Difference in Differences (DID), since the estimator, in its simplest form, is a difference between two differences.
In this section, we are going to look at identification using DID, estimation and estimation of sampling noise.
At first, we are going to assume that we have only access to two time periods.
In that case, estimation and inference are pretty straightforward. 
We will then examine the case of several time periods, but we will first allow for only one treatment date.
In that case, we will introduce the standard tools used by applied researchers to analyze these types of designs: the event study graph and the Two-Way Fixed Effects estimator (a.k.a. TWFE).
We will determine which effect is estimated by the TWFE estimator and what are the goals of the event study graph.
We will then look at the most complex case: the staggered design, where we have several time periods (strictly more than two) and the date of treatment differs across units.
In the staggered design, troubles start appearing for the TWFE estimator.
We will survey these problems and the proposed solutions to address them.
Finally, we will look at the combination of DID with instrumental variables (the DID-IV estimator) and see which specific types of problems happen there as well.
Let's get to it.

### Difference In Differences with two time periods

Before getting into the rigorous derivations, let's start with a very simple illustration using our workhorse example.

```{example}
How does DID perform and what does it look like in our example model?
```
Let's first generate a dataset with selection bias.

\begin{align*}
y_i^1 & = y_i^0+\bar{\alpha}+\theta\mu_i+\eta_i \\
y_i^0 & = \mu_i+\delta+U_i^0 \\
U_i^0 & = \rho U_i^B+\epsilon_i \\
y_i^B & =\mu_i+U_i^B \\
U_i^B & \sim\mathcal{N}(0,\sigma^2_{U}) \\
D_i   & = \uns{y_i^B+ V_i\leq\bar{y}} \\
V_i   & = \gamma(\mu_i-\bar{\mu}) + \omega_i \\
(\eta_i,\omega_i) & \sim\mathcal{N}(0,0,\sigma^2_{\eta},\sigma^2_{\omega},\rho_{\eta,\omega})
\end{align*}

```{r SimulDID,eval=TRUE,echo=FALSE,results='hide'}
set.seed(1234)
N <-1000
cov.eta.omega <- matrix(c(param["sigma2eta"],param["rhoetaomega"]*sqrt(param["sigma2eta"]*param["sigma2omega"]),param["rhoetaomega"]*sqrt(param["sigma2eta"]*param["sigma2omega"]),param["sigma2omega"]),ncol=2,nrow=2)
eta.omega <- as.data.frame(mvrnorm(N,c(0,0),cov.eta.omega))
colnames(eta.omega) <- c('eta','omega')
mu <- rnorm(N,param["barmu"],sqrt(param["sigma2mu"]))
UB <- rnorm(N,0,sqrt(param["sigma2U"]))
yB <- mu + UB 
YB <- exp(yB)
Ds <- rep(0,N)
V <- param["gamma"]*(mu-param["barmu"])+eta.omega$omega
Ds[yB+V<=log(param["barY"])] <- 1 
epsilon <- rnorm(N,0,sqrt(param["sigma2epsilon"]))
U0 <- param["rho"]*UB + epsilon
y0 <- mu +  U0 + param["delta"]
alpha <- param["baralpha"]+  param["theta"]*mu + eta.omega$eta
y1 <- y0+alpha
Y0 <- exp(y0)
Y1 <- exp(y1)
y <- y1*Ds+y0*(1-Ds)
Y <- Y1*Ds+Y0*(1-Ds)
```

Let's see how DID works on this data.

```{r PlotDID,eval=TRUE,fig.cap='Evolution of average outcomes in the treated and control group',fig.align='center',out.width='50%',echo=FALSE,results='hide',fig.pos='htbp'}
x <- c("Before","After")
y1.observed <- c(mean(yB[Ds==1]),mean(y[Ds==1]))
y1.counterfactual <- c(mean(yB[Ds==1]),mean(y0[Ds==1]))
y0.observed <- c(mean(yB[Ds==0]),mean(y0[Ds==0]))
y1.DID <- c(mean(yB[Ds==1]),mean(y0[Ds==0])+mean(yB[Ds==1])-mean(yB[Ds==0]))
data.DID.plot <- as.data.frame(c(y1.observed,y1.counterfactual,y0.observed,y1.DID))
colnames(data.DID.plot) <- c("Outcome") 
data.DID.plot$Period <- factor(rep(x,4),levels=c("Before","After"))
data.DID.plot$Group <- factor(c("Treated","Treated","Treated counterfactual","Treated counterfactual","Untreated","Untreated","Treated DID","Treated DID"),levels=c("Untreated","Treated","Treated counterfactual","Treated DID"))
data.DID.plot$Observed <- factor(c("Observed","Observed","Unobserved","Unobserved","Observed","Observed","Observed","Observed"),levels=c("Observed","Unobserved"))

WW.before <- mean(yB[Ds==1])-mean(yB[Ds==0])
WW.after <- mean(y[Ds==1])-mean(y[Ds==0])
BA.treated <- mean(y[Ds==1])-mean(yB[Ds==1])
BA.untreated <- mean(y[Ds==0])-mean(yB[Ds==0])
Counterfactual.after <- mean(yB[Ds==1])+BA.untreated
DID <- WW.after - WW.before 
TT <- mean(alpha[Ds==1])

ggplot(data.DID.plot,aes(x=Period,y=Outcome,group=Group,color=Group,shape=Group,linetype=Observed))+
    geom_line() +
    geom_point()+
    scale_linetype_discrete(guide='none') +
    theme_bw()
```

Figure \@ref(fig:PlotDID) shows the evolution of the mean log-outcomes for the treated and untreated groups over time in our simulated dataset.
We can see that in the **Before** period, outcomes ($y_i^B$ in that case) are much higher for the non participants than for the participants, in agreement with the selection rule that makes participation into the program more likely for individuals with lower pre-treatment outcomes.
The With/Without difference in outcomes before the program takes place is $\hat{\Delta}^{y^B}_{WW}=$ `r round(WW.before,3)`.
Second, we see that the difference between participants and non-participants decreases after receiving the treatment.
This is because the outcomes of the participants increase faster than the outcomes of the non participants.
As a consequence, the With/Without difference in outcomes after the program takes place is $\hat{\Delta}^{y}_{WW}=$ `r round(WW.after,3)`.
The DID estimator is built by comparing these two differences.
In our example, $\hat{\Delta}^{y}_{DID}=$ `r round(DID,3)`.
It is not too far from the true treatment effect of $\hat{\Delta}^{y}_{TT}=$ `r round(TT,3)`.

Figure \@ref(fig:PlotDID) also demonstrates that the DID estimator can also be seen as the difference between the Before/After differences in outcomes of the treated and the untreated.
The Before/After difference in outcomes for the non participants is $\hat{\Delta}^{y}_{BA|D=0}=$ `r round(BA.untreated,3)` while the Before/After difference for the participants is $\hat{\Delta}^{y}_{BA|D=1}=$ `r round(BA.treated,3)`, leading to the same DID estimand.
One way to understand the DID estimator is to see it as recreating the counterfactual trajectory of the participants (show as a discontinuous line on Figure \@ref(fig:PlotDID)) by using the trajectory of the non participants and making it start at the pre-treatment level of the participants.
This estimated counterfactual trajectory is shown as the purple continuous line at the bottom of Figure \@ref(fig:PlotDID).
In our example, the true counterfactual trajectory (the discontinuous line) and the estimated counterfactual trajectory almost coincide, making the estimated counterfactual outcome of the participants very close to their true counterfactual outcome (`r round(Counterfactual.after,3)` vs `r round(mean(y0[Ds==1]),3)`).
The difference between these two quantities measures the bias of the DID estimator, and we can see that it is very low in our example.
The fact that the Before/After difference in outcomes for the non participants approximates well the counterfactual Before/After difference in outcomes for the participants is **THE** crucial assumption of the DID estimator. 
It is called the parallel trends assumption.

#### Identification 

The formal setting for introducing the DID estimator is to start with two time periods, **Before** and **After** ($t=B$ and $t=A$ respectively).
Outcomes with and without the treatment in both periods are denoted $Y^d_{i,t}$, for $d\in\left\{0,1\right\}$ and $t\in\left\{B,A\right\}$.
Treatment participation in both periods is denoted $D_{i,t}$ for $t\in\left\{B,A\right\}$.
In the Before period, the treatment is unavailable, so that we get to observe the potential outcomes of the agents in the absence of the treatment.
These two very specific requirements of DID are encoded in the following way:

```{hypothesis,NoTreatmentFirst,name="No Treatment in the Before Period"}
We assume that no unit in the population receives the treatment in the Before period: $D_{i,B}=0$, $\forall i$.
```

Under Assumption \@ref(hyp:NoTreatmentFirst), and without loss of generality, we are going to write $D_i=D_{i,A}$.

```{hypothesis,NoAnticipationEffects,name="No Anticipation Effects"}
We assume that, in the Before period, agents cannot anticipate that the program will happen in the After period, or that they do not change their behavior as a consequence: $Y_{i,B}=Y^0_{i,B}$, $\forall i$.
```

A consequence of Assumptions \@ref(hyp:NoTreatmentFirst) and \@ref(hyp:NoAnticipationEffects) is that we can write observed outcomes as a function of treatment and potential outcomes using the usual switching equation:

\begin{align}\label{eqn:switchDID}
  Y_{i,t} & = Y^1_{i,t}D_{i,t} + Y^0_{i,t}(1-D_{i,t}).
\end{align}

The final very important assumption that we can make is to assume that the trends in the potential outcomes in the absence the treatment are the same for the treated and the untreated units:

```{hypothesis,ParallelTrends,name="Parallel Trends"}
We assume that the trends in the potential outcomes in the absence the treatment are the same for the treated and the untreated units:

\begin{align*}
    \esp{Y^0_{i,A}|D_i=1} - \esp{Y^0_{i,B}|D_i=1} & =   \esp{Y^0_{i,A}|D_i=0} - \esp{Y^0_{i,B}|D_i=0}.
\end{align*}
```

Assumption \@ref(hyp:ParallelTrends) is actually equivalent to assuming that selection bias is constant over time.
This is what this very simple lemma shows:

```{lemma,ParallelTrendsCstSelectionBias,name="Parallel Trends is Constant Selection Bias"}
Assumption \@ref(hyp:ParallelTrends) is equivalent to assuming that selection bias is constant over time:

\begin{align*}
    \esp{Y^0_{i,A}|D_i=1} - \esp{Y^0_{i,A}|D_i=0} & =   \esp{Y^0_{i,B}|D_i=1} - \esp{Y^0_{i,B}|D_i=0} .
\end{align*}
```

```{proof}
The proof follows immediately by adding $\esp{Y^0_{i,B}|D_i=1}-\esp{Y^0_{i,A}|D_i=0}$ to both sides of the equation in Assumption \@ref(hyp:ParallelTrends). 
```

Under these assumptions, we are ready to state the main identification result of this section:

```{theorem,IdentDID,name="DID identifies TT"}
Under Assumptions \@ref(hyp:NoTreatmentFirst), \@ref(hyp:NoAnticipationEffects) and \@ref(hyp:ParallelTrends), the DID estimator identifies the average effect of the Treatment on the Treated after the treatment:

\begin{align*}
    \Delta_{DID}^{Y} & =  \Delta^{Y_A}_{TT},
\end{align*}
```

with:

\begin{align*}
  \Delta^Y_{DID} & = \esp{Y_{i,A}|D_i=1} - \esp{Y_{i,B}|D_i=1} - (\esp{Y_{i,A}|D_i=0} - \esp{Y_{i,B}|D_i=0}),\\
  \Delta^{Y_A}_{TT} & = \esp{Y^1_{i,A}-Y^0_{i,A}|D_{i}=1}.
\end{align*}


```{proof}
\begin{align*}
  \Delta^Y_{DID} & = \esp{Y_{i,A}|D_i=1}-\esp{Y_{i,B}|D_i=1}-(\esp{Y_{i,A}|D_i=0}-\esp{Y_{i,B}|D_i=0}) \\
                & = \esp{Y^1_{i,A}|D_i=1}-\esp{Y^0_{i,B}|D_i=1}-(\esp{Y^0_{i,A}|D_i=0}-\esp{Y^0_{i,B}|D_i=0})\\
                & = \esp{Y^1_{i,A}|D_i=1}-\left(\esp{Y^0_{i,A}|D_i=0}+(\esp{Y^0_{i,B}|D_i=1}-\esp{Y^0_{i,B}|D_i=0})\right)
\end{align*}

where the second equality follows from Assumptions \@ref(hyp:NoTreatmentFirst) and \@ref(hyp:NoAnticipationEffects) and the switching equation, and the third equality follows from Lemma \@ref(lem:ParallelTrendsCstSelectionBias).
Under Assumption \@ref(hyp:ParallelTrends), we have:

\begin{align*}
  \esp{Y^0_{i,A}|D_i=1} & = \esp{Y^0_{i,A}|D_i=0} + (\esp{Y^0_{i,B}|D_i=1}-\esp{Y^0_{i,B}|D_i=0})
\end{align*}

As a consequence, we have:

\begin{align*}
  \Delta^Y_{DID} & = \esp{Y^1_{i,A}|D_i=1}-\esp{Y^0_{i,A}|D_i=1}\\
                & = \esp{Y^1_{i,A}-Y^0_{i,A}|D_i=1}\\
                & = \Delta^{Y_A}_{TT}.
\end{align*}
```

```{example}
How does the DID estimator behave in our example?
```
The Before/After comparison among the participants is equal to $\hat{\Delta}^{y}_{BA|D=1}=$ `r round(BA.treated,3)`.
The Before/After comparison among the non-participants is equal to $\hat{\Delta}^{y}_{BA|D=0}=$ `r round(BA.untreated,3)`.
The DID estimator is thus equal to $\hat{\Delta}^{y}_{DID}=\hat{\Delta}^{y}_{BA|D=1}-\hat{\Delta}^{y}_{BA|D=0}=$ `r round(BA.treated,3)` $-$ `r round(BA.untreated,3)` $=$ `r round(DID,3)`.
It is also equal to the difference between the before and after With/Without estimators.
The Before With/Without estimator is equal to $\hat{\Delta}^{y^B}_{WW}=$ `r round(WW.before,3)`.
The After With/Without estimator is equal to $\hat{\Delta}^{y}_{WW}=$ `r round(WW.after,3)`.
The DID estimator is thus equal to $\hat{\Delta}^{y}_{DID}=\hat{\Delta}^{y}_{WW}-\hat{\Delta}^{y^B}_{WW}=$ `r round(WW.after,3)` $-($ `r round(WW.before,3)` $)=$ `r round(DID,3)`.
This is not too far from the true effect of the treatment in the sample which is equal to $\hat{\Delta}^{y}_{TT}=$ `r round(TT,3)`.

Now, another very important question is whether the DID estimator is consistent, that is whether it is equal to $\Delta^{y}_{TT}$ in our model.
A necessary and sufficient condition for that is for the Parallel Trends Assumption \@ref(hyp:ParallelTrends) to hold. 
Indeed, it can be shown that the bias of the DID estimator is $\Delta^{y}_{B(DID)}=\Delta^{y}_{DID}-\Delta^{y}_{TT}=$ $\esp{y^0_{i}|D_i=1}  - \esp{y^B_{i}|D_i=1}-(\esp{y^0_{i}|D_i=0} - \esp{y^B_{i}|D_i=0})$.
Let us derive $\Delta^{y}_{B(DID)}$ in our example. 
Let us compute the trend in potential outcomes among the treated: 

\begin{align*}
  \esp{y^0_{i,A}|D_i=1} & - \esp{y^0_{i,B}|D_i=1} \\
  & = \esp{y^0_{i}|D_i=1} - \esp{y^B_{i}|D_i=1} \\
  & = \esp{\mu_i+\delta+U_i^0|D_i=1}-\esp{\mu_i+U_i^B|D_i=1} \\
  & = \esp{\mu_i|D_i=1}+\delta+\esp{U_i^0|D_i=1}\\
  & \phantom{=}-\esp{\mu_i|D_i=1}-\esp{U_i^B|D_i=1} \\
  & = \delta + \esp{\rho U_i^B+\epsilon_i|D_i=1}-\esp{U_i^B|D_i=1}\\
  & = \delta -(1-\rho)\esp{U_i^B|D_i=1}.
\end{align*}

Following the same line of reasoning, the trend in potential outcomes among the untreated is:

\begin{align*}
\esp{y^0_{i}|D_i=0} - \esp{y^B_{i}|D_i=0} & = \delta -(1-\rho)\esp{U_i^B|D_i=0}.
\end{align*}

As a consequence, the bias of the DID estimator in our model is:

\begin{align*}
  \Delta^{y}_{B(DID)} & = -(1-\rho)(\esp{U_i^B|D_i=1}-\esp{U_i^B|D_i=0}) \\
  & = -(1-\rho)(\esp{U_i^B|\mu_i+U_i^B+V_i\leq\bar{y}}-\esp{U_i^B|\mu_i+U_i^B+V_i>\bar{y}})
\end{align*}

Is this zero?
The answer actually is that it is not.
In order to see why, notice intuitively that the conditional expectation of $U_i^B$ is taken conditional on something correlated with $U_i^B$ being above or below some threshold.
As a consequence, the two values whose difference is taken in the parenthesis cannot be equal.
More formally, let us derive the formula for the bias of the DID estimator in our model, using the formula for the expectation of a truncated bivariate normal distribution:

\begin{align*}
  \Delta^{y}_{B(DID)}  & = -(1-\rho)(\esp{U_i^B|\mu_i+U_i^B+V_i\leq\bar{y}}-\esp{U_i^B|\mu_i+U_i^B+V_i>\bar{y}}) \\
                        & = (1-\rho)\left(\frac{\sigma^2_U}{(1+\gamma^2)\sigma^2_{\mu}+\sigma^2_U+\sigma^2_{\omega}}\right)
                                         \left(\frac{\phi\left(\frac{\bar{y}-\bar{\mu}}{(1+\gamma^2)\sigma^2_{\mu}+\sigma^2_U+\sigma^2_{\omega}}\right)}
                                            {\Phi\left(\frac{\bar{y}-\bar{\mu}}{(1+\gamma^2)\sigma^2_{\mu}+\sigma^2_U+\sigma^2_{\omega}}\right)}
                                            +\frac{\phi\left(\frac{\bar{y}-\bar{\mu}}{(1+\gamma^2)\sigma^2_{\mu}+\sigma^2_U+\sigma^2_{\omega}}\right)}
                                            {1-\Phi\left(\frac{\bar{y}-\bar{\mu}}{(1+\gamma^2)\sigma^2_{\mu}+\sigma^2_U+\sigma^2_{\omega}}\right)}
                                      \right)
\end{align*}

In order to compute the value of this parameter, and of the average treatment effect, we are going to use the package `tmtvnorm` which provides the moments from a truncated multivariate normal variable.
Here, we use the distribution of $(\alpha_i,U_i^B,\mu_i+U_i^B+V_i)$ which is normal with mean $(\bar{\alpha}+\theta\bar{\mu},0,\bar{\mu})$ and covariance matrix $\mathbf{D}$ with:

\begin{align*}
  \mathbf{D} &=  \left(\begin{array}{ccc}
                      \theta^2\sigma^2_{\mu}+ \sigma^2_{\eta} & 0 & (\theta+\gamma\theta)\sigma^2_{\mu}+\rho_{\eta,\omega}\sigma_{\eta}\sigma_{\omega}\\
                      0 & \sigma^2_U & \sigma^2_U \\
                      (\theta+\gamma\theta)\sigma^2_{\mu}+\rho_{\eta,\omega}\sigma_{\eta}\sigma_{\omega}& \sigma^2_U & (1+\gamma^2)\sigma^2_{\mu}+\sigma^2_U+\sigma^2_{\omega}
                      \end{array}\right)
\end{align*}

```{r BiasDIDPopulation,eval=TRUE,echo=TRUE,results='hide'}
mean.alpha.UB.yBV <- c(param['baralpha']+param['barmu']*param['theta'],0,param['barmu'])
cov.alpha.UB.yBV <- matrix(c(param['theta']^2*param['sigma2mu']+param['sigma2eta'],
                             0,
                             (param['theta']+param['gamma']*param['theta'])*param['sigma2mu']+param['rhoetaomega']*param['sigma2eta']*param['sigma2omega'],
                             0, 
                             param['sigma2U'],
                             param['sigma2U'],
                             (param['theta']+param['gamma']*param['theta'])*param['sigma2mu']+param['rhoetaomega']*param['sigma2eta']*param['sigma2omega'],
                             param['sigma2U'],
                            (1+param['gamma']^2)*param['sigma2mu']+param['sigma2U']+param['sigma2omega']),3,3,byrow=TRUE)

# cuts
#non participants
lower.cut.D0 <- c(-Inf,-Inf,log(param['barY']))
upper.cut.D0 <- c(Inf,Inf,Inf)
# participants
lower.cut.D1 <- c(-Inf,-Inf,-Inf)
upper.cut.D1 <- c(Inf,Inf,log(param['barY']))

# means 
TT <- mtmvnorm(mean=mean.alpha.UB.yBV,sigma=cov.alpha.UB.yBV,lower=lower.cut.D1,upper=upper.cut.D1,doComputeVariance=FALSE)[[1]][[1]]
mean.UB.D0 <- mtmvnorm(mean=mean.alpha.UB.yBV,sigma=cov.alpha.UB.yBV,lower=lower.cut.D0,upper=upper.cut.D0,doComputeVariance=FALSE)[[1]][[2]]
mean.UB.D1 <- mtmvnorm(mean=mean.alpha.UB.yBV,sigma=cov.alpha.UB.yBV,lower=lower.cut.D1,upper=upper.cut.D1,doComputeVariance=FALSE)[[1]][[2]]
B.DID <- -(1-param['rho'])*(mean.UB.D1-mean.UB.D0)
```

In our example, the population $TT$ is equal to $\Delta^y_{TT}=$ `r round(TT,3)`.
The DID estimator is equal to $\Delta^y_{DID}=$ `r round(TT+B.DID,3)`.
As a consequence, the bias of the DID estimator is equal to $\Delta^y_{B(DID)}=$ `r round(B.DID,3)`.

In order to make the DID estimator consistent for the $TT$ parameter, we need to impose that $\rho=1$.
When shocks are permanent, their bias remains constant over time and thus DID can estimate it without error. 
Let us generate new data that are compatible with that assumption.

```{r SimulDIDPT,eval=TRUE,echo=TRUE,results='hide'}
set.seed(1234)
N <-1000
param["rho"] <- 1
cov.eta.omega <- matrix(c(param["sigma2eta"],param["rhoetaomega"]*sqrt(param["sigma2eta"]*param["sigma2omega"]),param["rhoetaomega"]*sqrt(param["sigma2eta"]*param["sigma2omega"]),param["sigma2omega"]),ncol=2,nrow=2)
eta.omega <- as.data.frame(mvrnorm(N,c(0,0),cov.eta.omega))
colnames(eta.omega) <- c('eta','omega')
mu <- rnorm(N,param["barmu"],sqrt(param["sigma2mu"]))
UB <- rnorm(N,0,sqrt(param["sigma2U"]))
yB <- mu + UB 
YB <- exp(yB)
Ds <- rep(0,N)
V <- param["gamma"]*(mu-param["barmu"])+eta.omega$omega
Ds[yB+V<=log(param["barY"])] <- 1 
epsilon <- rnorm(N,0,sqrt(param["sigma2epsilon"]))
U0 <- param["rho"]*UB + epsilon
y0 <- mu +  U0 + param["delta"]
alpha <- param["baralpha"]+  param["theta"]*mu + eta.omega$eta
y1 <- y0+alpha
Y0 <- exp(y0)
Y1 <- exp(y1)
y <- y1*Ds+y0*(1-Ds)
Y <- Y1*Ds+Y0*(1-Ds)
```

Let's see how DID works on this data.

```{r PlotDIDPT,eval=TRUE,fig.cap='Evolution of average outcomes in the treated and control group when the Parallel Trends Assumption holds',fig.align='center',out.width='50%',echo=FALSE,results='hide',fig.pos='htbp'}
x <- c("Before","After")
y1.observed <- c(mean(yB[Ds==1]),mean(y[Ds==1]))
y1.counterfactual <- c(mean(yB[Ds==1]),mean(y0[Ds==1]))
y0.observed <- c(mean(yB[Ds==0]),mean(y0[Ds==0]))
y1.DID <- c(mean(yB[Ds==1]),mean(y0[Ds==0])+mean(yB[Ds==1])-mean(yB[Ds==0]))
data.DID.plot <- as.data.frame(c(y1.observed,y1.counterfactual,y0.observed,y1.DID))
colnames(data.DID.plot) <- c("Outcome") 
data.DID.plot$Period <- factor(rep(x,4),levels=c("Before","After"))
data.DID.plot$Group <- factor(c("Treated","Treated","Treated counterfactual","Treated counterfactual","Untreated","Untreated","Treated DID","Treated DID"),levels=c("Untreated","Treated","Treated counterfactual","Treated DID"))
data.DID.plot$Observed <- factor(c("Observed","Observed","Unobserved","Unobserved","Observed","Observed","Observed","Observed"),levels=c("Observed","Unobserved"))

WW.before <- mean(yB[Ds==1])-mean(yB[Ds==0])
WW.after <- mean(y[Ds==1])-mean(y[Ds==0])
BA.treated <- mean(y[Ds==1])-mean(yB[Ds==1])
BA.untreated <- mean(y[Ds==0])-mean(yB[Ds==0])
Counterfactual.after <- mean(yB[Ds==1])+BA.untreated
DID <- WW.after - WW.before 
TT <- mean(alpha[Ds==1])

ggplot(data.DID.plot,aes(x=Period,y=Outcome,group=Group,color=Group,shape=Group,linetype=Observed))+
    geom_line() +
    geom_point()+
    scale_linetype_discrete(guide='none') +
    theme_bw()
```

Now, the counterfactual change in outcome for the treated and its approximation using the trend experienced by the untreated are extremely close, as the curves *Treated counterfactual* and *Treated DID* show on Figure \@ref(fig:PlotDIDPT).

#### Estimation

Estimation of $TT$ under the $DID$ assumptions can be performed in a variety of ways: using directly the DID formula, using OLS with group fixed effects, using OLS with individual and time dummy variables, using first differences and using the within transformation (also known as the Two-Way Fixed Effects or TWFE estimator).
With only two periods of data and a fully balanced panel, all of these estimators are actually numerically equivalent. 
Let's examine them in turn.

##### Using the DID formula

One could go directly and use the DID formula of Theorem \@ref(thm:IdentDID).
The sample DID estimator is thus equal to:

\begin{align*} 
  \hat{\Delta}^Y_{DID} & = \frac{\sum_{i=1}^NY_{i,A}D_i}{\sum_{i=1}^ND_i} -\frac{\sum_{i=1}^NY_{i,B}D_i}{\sum_{i=1}^ND_i} - \left(\frac{\sum_{i=1}^NY_{i,A}(1-D_i)}{\sum_{i=1}^N(1-D_i)} -\frac{\sum_{i=1}^NY_{i,B}(1-D_i)}{\sum_{i=1}^N(1-D_i)}\right).
\end{align*}

```{example}
In our example, let's see how this estimator works.
```
The Before/After comparison among the participants is equal to $\hat{\Delta}^{y}_{BA|D=1}=$ `r round(BA.treated,3)`.
The Before/After comparison among the non-participants is equal to $\hat{\Delta}^{y}_{BA|D=0}=$ `r round(BA.untreated,3)`.
The DID estimator is thus equal to $\hat{\Delta}^{y}_{DID}=\hat{\Delta}^{y}_{BA|D=1}-\hat{\Delta}^{y}_{BA|D=0}=$ `r round(BA.treated,3)` $-$ `r round(BA.untreated,3)` $=$ `r round(DID,3)`.
It is also equal to the difference between the before and after With/Without estimators.
The Before With/Without estimator is equal to $\hat{\Delta}^{y^B}_{WW}=$ `r round(WW.before,3)`.
The After With/Without estimator is equal to $\hat{\Delta}^{y}_{WW}=$ `r round(WW.after,3)`.
The DID estimator is thus equal to $\hat{\Delta}^{y}_{DID}=\hat{\Delta}^{y}_{WW}-\hat{\Delta}^{y^B}_{WW}=$ `r round(WW.after,3)` $-($ `r round(WW.before,3)` $)=$ `r round(DID,3)`.
This is not too far from the true effect of the treatment in the sample which is equal to $\hat{\Delta}^{y}_{TT}=$ `r round(TT,3)`.
In the population, the $TT$ parameter has not changed, since its computation does not involve $\rho$.
We still have $\Delta^y_{TT}=$ `r round(TT,3)`.

##### Using the Least Squares pooling DID estimator

The most basic regression-based way to implement DID is to run a linear regression of outcomes on a treatment group dummy, a time dummy and their interaction.
The interaction captures the effect of the treatment estimated using DID.
The way it works is as follows: estimate the following equation using OLS and use $\hat{\beta}_{OLS}$ as your DID estimate: $\hat{\beta}_{OLS}=\hat{\Delta}^{Y}_{DID}$. 
\begin{align*}
  	Y_i &  = \alpha +  \mu D_i + \delta T_i + \beta D_iT_i + \epsilon_i.
\end{align*}
$D_i$ is our usual treatment indicator while $T_i$ takes value one when observation $i$ is observed in the *After* and zero otherwise.

```{example}
Let's see how this works in our example.
```
Before estimating the model, we need to build a data frame with all the necessary variables.

```{r RegDID,eval=TRUE,echo=TRUE,results='hide'}
# building a data frame
data.DID <- as.data.frame(cbind(c(y,yB),c(Ds,Ds),c(rep(1,N),rep(0,N))))
colnames(data.DID) <- c('y','D','T')

# running the OLS regression
reg.DID <- lm(y ~ D + T + D*T,data = data.DID)

# coefficients
yB.D0.reg <-  coef(reg.DID)[[1]]
WW.before.reg <-  coef(reg.DID)[[2]]
BA.untreated.reg <- coef(reg.DID)[[3]]
DID.est.reg <- coef(reg.DID)[[4]]

# comparisons
yB.D0.sample <- mean(yB[Ds==0])
```

The estimate of $\hat{\beta}_{OLS}$ in our sample is equal to `r round(DID.est.reg,3)`.
It is exactly equal to $\hat{\Delta}^{y}_{DID}$ as estimated just above.
What is interesting with the regression-based DID approach is that the other coefficients in the regression have a direct interpretation.
For example, the constant $\alpha$ estimates the mean outcome in the untreated group before the treatment.
In our case, we have $\hat{\alpha}_OLS=$ `r round(yB.D0.reg,3)`.
Remember that in our sample, the average outcomes of the untreated before the treatment is equal to $\hatesp{y_i^B|D_i=0}=$ `r round(yB.D0.sample,3)`.
$\mu$, the coefficient in front of the $D_i$ dummy, estimates the With/Without estimator before the treatment.
In our case, we have $\hat{\mu}_{OLS}=$ `r round(WW.before.reg,3)`.
Remember that in our sample, the With/Without estimator before the treatment is equal to $\hat{\Delta}^{y^B}_{WW}=$ `r round(WW.before,3)`.
$\delta$, the coefficient in front of the $T_i$ dummy, estimates the Before/After change in outcomes among the untreated.
In our case, we have $\hat{\delta}_{OLS}=$ `r round(BA.untreated.reg,3)`.
Remember that in our sample, the Before/After estimator among the untreated is equal to $\hat{\Delta}^{y}_{BA|D=0}=$ `r round(BA.untreated,3)`.

```{remark}
A pretty cool property of the regression-based DID estimator is that is does not require panel data.
It works even with repeated cross sections, *i.e.* when observations are drawn from the same population in both periods but are not the same.
```

##### Using First Differences

In the presence of panel data, an alternative to the regression-based DID estimator is the first-difference estimator. 
It simply regresses the change over time in outcomes on the treatment dummy:
\begin{align*}
  	Y_{i,A}-Y_{i,B} &  = \alpha^{FD} +  \beta^{FD} D_i + \epsilon^{FD}_i.
\end{align*}
The coefficient $\beta^{FD}$ estimated by OLS is an estimate of the DID parameter.

```{example}
Let's see how this works in our example.
```
Before running the model, we need to generate first the differenced estimates.
One very simple way to do that is simply to take the difference between the before and the after outcome vectors.

```{r RegFD,eval=TRUE,echo=TRUE,results='hide'}
# building a data frame
data.FD <- as.data.frame(cbind(y-yB,Ds))
colnames(data.FD) <- c('BAy','D')

# running the OLS regression
reg.FD <- lm(BAy ~ D,data = data.FD)

# coefficients
BA.untreated.FD <- coef(reg.FD)[[1]]
DID.est.FD <- coef(reg.FD)[[2]]
```

The estimate of $\hat{\beta}^{FD}_{OLS}$ in our sample is equal to `r round(DID.est.FD,3)`.
It is exactly equal to $\hat{\Delta}^{y}_{DID}$ as estimated just above.
Note also that $\alpha^{FD}$ estimates the Before/After change in outcomes among the untreated.
In our case, we have $\hat{\alpha}^{FD}_{OLS}=$ `r round(BA.untreated.FD,3)`.
Remember that in our sample, the Before/After estimator among the untreated is equal to $\hat{\Delta}^{y}_{BA|D=0}=$ `r round(BA.untreated,3)`.

##### Using the Least Squares Dummy Variables estimator

One very computer-intensive way to estimate $TT$ in a DID setting is to use the OLS estimator supplemented with dummies for each observation and for each time period, also called the Least-Squares Dummy Variables estimator.
In practice, the estimator is based on the following regression:
\begin{align*}
  	Y_{i,t} &  = \sum_{j=1}^N\mu_j\uns{j=i} + \sum_{l=0}^1\delta_l\uns{l=t} + \beta^{LSDV} D_{i,t} + \epsilon^{LSDV}_{i,t}.
\end{align*}
The notation is generally simplified as follows:
\begin{align*}
  	Y_{i,t} &  = \mu_i + \delta_t + \beta^{TWFE} D_{i,t} + \epsilon^{TWFE}_{i,t},
\end{align*}

This last estimator is generally called the Two-Way Fixed Effects estimator, since it has two-sets of so-called fixed effects (individual fixed effects, $\mu_i$, and time fixed effects $\delta_t$).
I will write it using this second, more compact, formulation, but I think the first formulation encapsulates better how the Least-Squares Dummy Variables estimator works.
In what follows, we will see other ways of estimating the Two-Way Fixed Effects model, but for now, let us focus on the Least-Squares Dummy Variables estimator. 
The way it works is simply by throwing a bunch of dummy variables in the regression.

```{example}
Let's see how the Least Squares Dummy Variable works in our example.
For that, we need to generate one dummy variable for each individual $i$ in our sample.
This is made simple by the `factor` function in `R`. 
We are also going to run the model without a constant, so that all the fixed effects are identified.
```

```{r RegLSDV,eval=TRUE,echo=TRUE,results='hide'}
# adding one column to the DID data frame with the individual index for each observation of the same $i$
data.DID$indiv <- as.factor(c(1:N,1:N))
# generating Dit (time varying)
data.DID$Dit <- data.DID$D*data.DID$T
# running the LSDV estimator
reg.LSDV <- lm(y~-1 + Dit + as.factor(T) + indiv,data=data.DID)
# result
DID.est.LSDV <- coef(reg.LSDV)[[1]]
```

The Least-Squares Dummy Variables estimate of $TT$ is equal to: $\hat{\beta}^{LSDV}=$ `r round(DID.est.LSDV,3)`.

```{remark}
The term *fixed effect* is specific to the panel data literature in econometrics. 
It refers to the fact that both $\mu_i$ and $\delta_t$ are allowed to be correlated with $D_{i,t}$ in this model.
This is in contrast to the *random effects model* where $\mu_i$ and $\delta_t$ are assumed to be independent of the regressors of interest.
```

##### Using the Within estimator

You might have noticed that the Least-Squares Dummy Variables estimator took some time to compute on your computer. 
This is because it requires the inversion of a very large matrix, as large as the number of fixed effects plus one. 
The size of this computation increases as the number of observation and time periods increases, meaning that this computation might become practically unfeasible in very large datasets. 
Several tricks have been developed to decrease the computational burden of the estimation of the Two-Way Fixed Effects model. 
One approach is to use the First Difference estimator.
Another approach is the Within estimator. 
The way the Within estimator works is by taking the difference between each observation and its mean over time or over individuals. 
More precisely, the Within estimator estimates the following model by OLS:
\begin{align*}
  	Y_{i,t}-\frac{1}{2}\sum_{t=0}^1Y_{i,t} &  = \delta^{W}_t + \beta^{W}(D_{i,t}-\frac{1}{2}\sum_{t=0}^1D_{i,t}) + \epsilon^{W}_{i,t}.
\end{align*}

The reason why this trick works is because of the shape of the Two-Way Fixed Effects model.
Indeed, taking the average of the Two-Way Fixed Effects model over time gives:
\begin{align*}
  	\frac{1}{2}\sum_{t=0}^1Y_{i,t} &  = \mu_i + \frac{1}{2}\sum_{t=0}^1\delta_t + \beta^{TWFE}\frac{1}{2}\sum_{t=0}^1D_{i,t} + \frac{1}{2}\sum_{t=0}^1\epsilon^{TWFE}_{i,t}.
\end{align*}
Taking the difference between the Two-Way Fixed Effects model and its time-averaged version gives the Within estimator.
The key is that the differencing gets rid of the individual fixed effects parameter $\mu_i$ and thus makes it unnecessary to estimate it.
The set of parameters to estimate is thus much smaller than in the Least-Squares Dummy Variables estimator.

```{example}
Let's see how the Within estimator works in our example.
For that, we need to compute the average over time of the outcome and of the treatment for each observation in our dataset. 
This is made simple by the `summarize` function of the `dplyr` package.
```

```{r RegW,eval=TRUE,echo=TRUE,results='hide'}
# generating the time means of Y and Dit
TimeMeansYDit <- data.DID %>%
                  group_by(indiv) %>%
                  summarize(
                    TimeMeanY = mean(y),
                    TimeMeanDit = mean(Dit)
                  )
# doubling the observations to be able to take the difference in both periods
TimeMeansYDit <- rbind(TimeMeansYDit,TimeMeansYDit)                  
# taking the difference in both periods
data.DID$W.y <- data.DID$y-TimeMeansYDit$TimeMeanY
data.DID$W.Dit <- data.DID$Dit-TimeMeansYDit$TimeMeanDit
# running the within estimator
reg.W <- lm(W.y~-1 + W.Dit + as.factor(T),data=data.DID)
# result
DID.est.W <- coef(reg.W)[[1]]
```

The Within estimate of $TT$ is equal to: $\hat{\beta}^{W}=$ `r round(DID.est.W,3)`.

The `plm` package directly implements the Within transformation.
The same package also estimates the First Difference model and the Least Squares pooling DID estimator.
Let's see how this works.

```{r RegWplm,eval=TRUE,echo=TRUE,results='hide'}
# running the within estimator
reg.W.plm <- plm(y ~  Dit + as.factor(T) , data = data.DID, index= c("indiv", "T"), model = "within")
# result
DID.est.W.plm <- coef(reg.W.plm)[[1]]

# running the first difference estimator
reg.FD.plm <- plm(y ~  Dit + as.factor(T) , data = data.DID, index= c("indiv", "T"), model = "fd")
# result
DID.est.FD.plm <- coef(reg.FD.plm)[[2]]

# running the OLS pooling DID estimator
reg.OLS.plm <- plm(y ~ as.factor(T) + D + Dit , data = data.DID, index= c("indiv", "T"), model = "pooling")
# result
DID.est.OLS.plm <- coef(reg.OLS.plm)[[4]]
```

As expected, `plm` gives the following estimates for $TT$: $\hat{\beta}^{W}=$ `r round(DID.est.W.plm,3)`, $\hat{\beta}^{FD}=$ `r round(DID.est.FD.plm,3)` and $\hat{\beta}^{OLS}=$ `r round(DID.est.OLS.plm,3)`. 

##### Using fast estimators of the Two-Way Fixed Effects model

All the estimators of the TWFE model that we have seen so far have issues. 
The OLS pooling DID estimator does not account for the panel structure of the data when it exists.
It does not alter the precision of the estimator but it makes it mode difficult to account for more dimensions of fixed effects than two.
The First Difference estimator, similarly, cannot easily account for more than two sets of fixed effects.
The Least Squares Dummy variable is slow because of the very large matrix inversion problem.
Therefore, applied econometricians tend to prefer using the Within estimator in practice.
The Within estimtor of the Two-Way Fixed Effects model is not without problems as well.
As the sample size grows large, or the number of fixed effects increases, it becomes more and more difficult to compute the within transformation.
As a consequence, recent packages have proposed to optimize the computation of the TWFE model using various computational tricks. 
Let's examine two in turn.

###### The Alternating Projections method

The `lfe` package in `R` implements an alternating projections method to estimate the $N$-Way Fixed effects model.
It is based on an algorithm proposed by [Gaure (2013)](https://doi.org/10.1016/j.csda.2013.03.024). 
The basic idea of [Gaure (2013)](https://www.econstor.eu/bitstream/10419/47280/1/637363027.pdf) is to repeat centering on the means of the fixed effects (the within operation) in an alternating manner between the various fixed effects dimensions until convergence. 

```{example}
Let's see how the `lfe` estimator works in our example.
```

```{r RegWlfe,eval=TRUE,echo=TRUE,results='hide'}
# running the within estimator
reg.W.lfe <- felm(y ~  Dit + as.factor(T) | indiv , data = data.DID)
# result
DID.est.W.lfe <- coef(reg.W.lfe)[[1]]
```

As expected, `lfe` gives the following estimate for $TT$: $\hat{\beta}^{AP}=$ `r round(DID.est.W.lfe,3)`.

###### The Likelihood Concentration method

One problem with the `lfe` package is that it works only for linear models.
The [`fixest`](https://cran.r-project.org/web/packages/fixest/vignettes/fixest_walkthrough.html) package in `R` proposes a solution for estimating fixed effects models in non-linear cases as well.
The solution is based on the concentrated likelihood as explained in [Berge (2018)](https://wwwen.uni.lu/content/download/110162/1299525/file/2018_13).
The intuition is as follows.
We first postulate a value for the treatment effect and the coefficient on the time dummies and we estimate each of the individual fixed effects using maximum likelihood.
We then use maximum likelihood to find the treatment effect using the values of the fixed effects estimated in the previous step. 
This seems complicated but the key idea is to separate the estimation of the fixed effects from the estimation of the parameters of interest. 

```{example}
Let's see how the `fixest` estimator works in our example.
```

```{r RegWfixest,eval=TRUE,echo=TRUE,results='hide'}
# running the within estimator
reg.W.fixest <- feols(y ~  Dit + as.factor(T) | indiv , data = data.DID)
# result
DID.est.W.fixest <- coef(reg.W.fixest)[[1]]
```

As expected, `fixest` gives the following estimate for $TT$: $\hat{\beta}^{LC}=$ `r round(DID.est.W.fixest,3)`.

##### Equivalence between the various DID methods with two time periods

The above results suggest that all DID estimators are equivalent when working with two time periods.
The following theorem actually states this result rigorously:

```{theorem,EstimDID,name="All DID estimators are numerically equivalent with two time periods"}
Under Assumptions \@ref(hyp:NoTreatmentFirst), \@ref(hyp:NoAnticipationEffects) and \@ref(hyp:ParallelTrends), in a panel with only two periods of data, all the DID estimators are numerically equivalent: $\hat{\beta}^{OLS}=\hat{\beta}^{FD}=\hat{\beta}^{W}=\hat{\beta}^{LSDV}=\hat{\beta}^{AP}=\hat{\beta}^{LC}=\hat{\Delta}^Y_{DID}$. 
```

```{proof}
See Section \@ref(proofEstimDID).
```

Finally, let's see how our estimator varies across sampling replications.
A key difference is whether we have access to panel data or not. 
Indeed, estimates from a repeated cross section are going to be more noisy since they are going to sample different people in different periods and thus are going to be affected by sampling noise stemming from the fixed effects. 
This is not going to be the case with panel data, since all the estimators based on the TWFE estimator differentiate out the individual fixed effects.

**Do the simulations**

#### Inference

**Derive the distribution of the estimators with repeated cross sections and panels**

### Differences In Differences with more than two-time periods but only one treatment date

In real life, we generally have access to several time periods before and after the treatment date.
What happens to DID in that case?
The first thing is that, with several periods after the treatment date, we now have a time-varying $TT$: $\Delta^{TT}_{Y_t}=\esp{Y^1_{i,t}-Y^0_{i,t}|D_i=1}$.

**Event Study graph and period specific treatment effects**

**What are the weights for the TWFE estimator?**
*Should be derived from the staggered case*


### Differences In Differences with a staggered design

### Difference In Differences with Instrumental Variables

