# Mediation Analysis 

```{r librariesBis,echo=FALSE,message=FALSE,error=FALSE,warning=FALSE}
library(MASS)
library(RColorBrewer)
library(gplots)
library(ggplot2)
library(snowfall)
library(cowplot)
library(sandwich)
library(ggpubr)
library(dplyr)
library(metafor)
library(stats)
library(tmvtnorm)
library(AER)
library(stargazer)
library(plm)
library(lfe)
library(fixest)
library(tidyr)
library(purrr)
library(stringr)
library(did)
library(DIDmultiplegt)
library(didimputation)
library(did2s)
library(recipes)
library(tidyverse)
library(ggthemes)
library(fastDummies)
library(ggdag)
```

When we have estimated the treatment effect of a program, we sometimes wonder by which channels the program impact has been obtained.
For example, has a Job Training Program been successful because it has increased the human capital of an agent, or simply by signalling to employers her motivation?
The question of separating between the various channels into which a program impact can be decomposed becomes especially important when a program has several components, and we wish to ascertain which one is the more important. 
Another reason why we might be interested in which channel precisely is responsible for the program impact is because which channel dominates might give us indications about which theoretical mechanism is at play.

In this chapter, I am going to first delineate the general framework for mediation analysis and the way mediation analysis can be undertaken in the ideal case of a Randomized Controlled Trial.
Then, I am going to present the fundamental problem of mediation analysis (which turns out to be one version of the confounders problem we know all too well) and the various techniques that have been developed in order to solve for it.

## Mediation analysis: a framework

Mediation analysis posits the existence of a mediator, $M_i$, which is driving part or the totality of the effect of the treatment on outcome $Y_i$. 
One way to understand how a mediator works is to draw a Direct Acyclic Graph or DAG which represents the relationship between the variables.
The `ggdag` package helps to draw a DAG easily.
Let's give it a try:

```{r DAGMediation,eval=TRUE,echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,results='hide',fig.cap='DAG of the impact of D on Y partially mediated through M',fig.align='center',out.width='50%',fig.pos='htbp'}
# structure of DAG
our_dag <- ggdag::dagify(M ~ D,
                         Y ~ D,
                         Y ~ M,
                         coords=list(x = c(D = 0, M = 1, Y = 2),
                                     y = c(D = 0, M = 1, Y = 0))
                        )
# plotting DAG
ggdag::ggdag(our_dag) + theme_void()
```

We are first going to define mediated and unmediated treatment effects and see how they work in our example and we'll then try to understand better how exactly mediation works and the mechanisms that are behind these treatment effects.

### Defining mediated and unmediated treatment effects

Let's start with a binary mediator, in order to keep things simple: $M_i\in\left\{0,1\right\}$.
We can define four potential outcomes $Y_i^{d,m}$, $(d,m)\in\left\{0,1\right\}^2$.
We can also define two potential outcomes for the mediator: $M^d_i$, $d\in\left\{0,1\right\}$.

The switching equation can now be written as follows:

\begin{align*}
  Y_i & = 
    \begin{cases}
    Y_i^{1,1} & \text{if } D_i=1 \text{ and }M_i=1\\
    Y_i^{1,0} & \text{if } D_i=1 \text{ and }M_i=0\\
    Y_i^{0,1} & \text{if } D_i=0 \text{ and }M_i=1\\
    Y_i^{0,0} & \text{if } D_i=0 \text{ and }M_i=0
    \end{cases} \\
    & = Y_i^{1,1}D_iM_i + Y_i^{0,1}(1-D_i)M_i  + Y_i^{1,0}D_i(1-M_i)+ Y_i^{0,0}(1-D_i)(1-M_i).
\end{align*}

We are now equipped to define two sets of causal mediation effects: the mediated (or indirect) effect and the unmediated (or direct) effect (we follow [Imai, Keene and Tingley (2010)](https://imai.fas.harvard.edu/research/files/BaronKenny.pdf)):

\begin{align*}
  \Delta^{Y_{m(d)}}_{i} & = Y_i^{d,M_i^1}-Y_i^{d,M_i^0}\\
  \Delta^{Y_{u(d)}}_{i} & = Y_i^{1,M_i^d}-Y_i^{0,M_i^d}.
\end{align*}

$\Delta^{Y_{m(d)}}_{i}$ is the causal effect of the treatment on the outcome acting through the mediator only, while keeping the direct effect of the treatment fixed at $d$.
$\Delta^{Y_{u(d)}}_{i}$ is the causal effect of the treatment on the outcome acting through the treatment only, while keeping the value of the mediator fixed at $M_i(d)$.
In the absence of interaction effects between the treatment and the mediator, we have $\Delta^{Y_{m(0)}}_{i}=\Delta^{Y_{m(1)}}_{i}=\Delta^{Y_{m}}_{i}$ and $\Delta^{Y_{u(0)}}_{i}=\Delta^{Y_{u(1)}}_{i}=\Delta^{Y_{u}}_{i}$.

The individual effect of the treatment can be decomposed in two components:

\begin{align*}
  \Delta^Y_{i} & = Y_i^{1,M_i^1}-Y_i^{0,M_i^0}\\
                 & = Y_i^{1,M_i^1}-Y_i^{1,M_i^0}+Y_i^{1,M_i^0}-Y_i^{0,M_i^0}=\Delta^{Y_{m(1)}}_{i}+\Delta^{Y_{u(0)}}_{i}\\
                 & = Y_i^{1,M_i^1}-Y_i^{0,M_i^1}+Y_i^{0,M_i^1}-Y_i^{0,M_i^0}=\Delta^{Y_{u(1)}}_{i}+\Delta^{Y_{m(0)}}_{i}.
\end{align*}

In the absence of interaction effects between the mediator and the treatment, this decomposition is unique.
The decomposition can also be applied to the TT parameter:

\begin{align*}
  \Delta^Y_{TT} & = \esp{Y_i^{1,M_i^1}-Y_i^{0,M_i^0}|D_i=1}\\
             & = \Delta^{Y_{m(1)}}_{TT}+\Delta^{Y_{u(0)}}_{TT}\\
            & = \Delta^{Y_{u(1)}}_{TT}+\Delta^{Y_{m(0)}}_{TT},
\end{align*}

with:

\begin{align*}
  \Delta^{Y_{m(d)}}_{TT} & = \esp{\Delta^{Y_{m(d)}}_{i}|D_i=1}\\
  \Delta^{Y_{u(d)}}_{TT} & = \esp{\Delta^{Y_{u(d)}}_{i}|D_i=1}.
\end{align*}

Here again, the decomposition will be unique if there are no interactions between the treatment and the mediator variable. 

```{example}
Let's see how this works in our example.
```

The first order of business is to set up a model:

\begin{align*}
    y_i^{1,1} & = y_i^0+\bar{\alpha}+ \tau_1 +\theta\mu_i+\eta_i \\
    y_i^{1,0} & = y_i^0+\bar{\alpha}+\theta\mu_i+\eta_i \\
    y_i^{0,1} & = \mu_i+\delta + \tau_0+U_i^0 \\
    y_i^{0,0} & = \mu_i+\delta+U_i^0 \\
    U_i^0 & = \rho U_i^B+\epsilon_i \\
    y_i^B & =\mu_i+U_i^B \\
    U_i^B & \sim\mathcal{N}(0,\sigma^2_{U}) \\
    M_i   & = \uns{\xi y_i^B + \psi D_i+ V^M_i\leq\bar{y}_M} \\
    V^M_i & = \gamma_M(\mu_i-\bar{\mu}) + \omega^M_i \\
    D_i   & = \uns{y_i^B+ V_i\leq\bar{y}} \\
    V_i   & = \gamma(\mu_i-\bar{\mu}) + \omega_i \\
    (\eta_i,\omega_i,\omega^M_i) & \sim\mathcal{N}(0,0,0,\sigma^2_{\eta},\sigma^2_{\omega},\sigma^2_{\omega_M},\rho_{\eta,\omega},\rho_{\eta,\omega_M},\rho_{\omega,\omega_M}) 
\end{align*}

Let us choose some parameter values and simulate the model.

```{r param.Mediation,eval=TRUE,echo=TRUE,results='hide'}
param <- c(8,.5,.28,1500,1500,0.9,0.01,0.05,0.05,0.05,0.1,0.2,0.1,1,-0.25,0.1,0.05,7.98,0.28,1,0,0,0)
names(param) <- c("barmu","sigma2mu","sigma2U","barY","barYM","rho","theta","sigma2epsilon","sigma2eta","delta","baralpha","tau1","tau0","xi","psi","gamma","gammaM","baryB","sigma2omega","sigma2omegaM","rhoetaomega","rhoetaomegaM","rhoomegaomegaM")
```

Let us now simulate the data:

```{r simul.Mediation,eval=TRUE,echo=TRUE,results='hide'}
set.seed(1234)
N <-1000
cov.eta.omega.omegaM <- matrix(c(param["sigma2eta"],param["rhoetaomega"]*sqrt(param["sigma2eta"]*param["sigma2omega"]),param["rhoetaomegaM"]*sqrt(param["sigma2eta"]*param["sigma2omegaM"]),
                                 param["rhoetaomega"]*sqrt(param["sigma2eta"]*param["sigma2omega"]),param["sigma2omega"],param["rhoomegaomegaM"]*sqrt(param["sigma2omega"]*param["sigma2omegaM"]),
                                 param["rhoetaomegaM"]*sqrt(param["sigma2eta"]*param["sigma2omegaM"]),param["rhoomegaomegaM"]*sqrt(param["sigma2omega"]*param["sigma2omegaM"]),param["sigma2omegaM"]),ncol=3,nrow=3)
eta.omega <- as.data.frame(mvrnorm(N,c(0,0,0),cov.eta.omega.omegaM))
colnames(eta.omega) <- c('eta','omega','omegaM')
mu <- rnorm(N,param["barmu"],sqrt(param["sigma2mu"]))
UB <- rnorm(N,0,sqrt(param["sigma2U"]))
yB <- mu + UB 
YB <- exp(yB)
Ds <- rep(0,N)
V <- param["gamma"]*(mu-param["barmu"])+eta.omega$omega
Ds[yB+V<=log(param["barY"])] <- 1 
VM <- param["gammaM"]*(mu-param["barmu"])+eta.omega$omegaM
M <- rep(0,N)
M[param['xi']*yB+param['psi']*Ds+VM<=log(param["barYM"])] <- 1 
M1 <- rep(0,N)
M1[param['xi']*yB+param['psi']+VM<=log(param["barYM"])] <- 1 
M0 <- rep(0,N)
M0[param['xi']*yB+VM<=log(param["barYM"])] <- 1 
epsilon <- rnorm(N,0,sqrt(param["sigma2epsilon"]))
U0 <- param["rho"]*UB + epsilon
alpha <- param["baralpha"]+  param["theta"]*mu + eta.omega$eta
y00 <- mu +  U0 + param["delta"]
y01 <- mu +  U0 + param['tau0']+ param["delta"]
y10 <- y00+alpha
y11 <- y00+alpha+param['tau1']
y1 <- y11*M1+y10*(1-M1)
y0 <- y01*M0+y00*(1-M0)
y1M1 <- y00+alpha+param['tau1']*M1
y1M0 <- y00+alpha+param['tau1']*M0
y0M1 <- y00 + param['tau0']*M1
y0M0 <- y00 + param['tau0']*M0
y <- y11*Ds*M+y10*Ds*(1-M)+y01*(1-Ds)*M+y00*(1-Ds)*(1-M)
```

Let us finally compute the values of TT and of the mediated and unmediated average treatment effects on the treated in the sample.

```{r simul.Mediation.TT,eval=TRUE,echo=TRUE,results='hide'}
# treatment on the treated
TT <- mean(y1[Ds==1]-y0[Ds==1])
# mediated treatment effects
TTm1 <- mean(y1M1[Ds==1]-y1M0[Ds==1])
TTm0 <- mean(y0M1[Ds==1]-y0M0[Ds==1])
# unmediated treatment effects
TTu1 <- mean(y1M1[Ds==1]-y0M1[Ds==1])
TTu0 <- mean(y1M0[Ds==1]-y0M0[Ds==1])

# preparing graph
mediation.example <- data.frame(TT=c(TTm1,TTu0,TTm0,TTu1),Effect=c("TTm1","TTu0","TTm0","TTu1"),Type=c("Mediated","Unmediated","Mediated","Unmediated"),Decomposition=c("m1u0","m1u0","m0u1","m0u1")) %>%
                      mutate(
                        Effect=factor(Effect,levels=c("TTm1","TTu0","TTm0","TTu1")),
                        Type=factor(Type,levels=c("Mediated","Unmediated")),
                        Decomposition=factor(Decomposition,levels=c("m1u0","m0u1"))
                      )
```

The average effect of the treatment on the treated is equal to `r round(TT,3)`.
It can be decomposed in two ways.
First, the impact mediated through the mediator $M_i$ when $D_i$ is fixed at 1 (`r round(TTm1,3)`) and the effect not mediated through $M_i$ (with $M_i$ fixed at $M_i^0$) (`r round(TTu0,3)`).
Second, the impact mediated through the mediator $M_i$ when $D_i$ is fixed at 0 (`r round(TTm0,3)`) and the effect not mediated through $M_i$ (with $M_i$ fixed at $M_i^1$) (`r round(TTu1,3)`).
Let us plot the results:

```{r PlotMediationExample,eval=TRUE,warning=FALSE,error=FALSE,message=FALSE,echo=TRUE,fig.cap="Decomposition of the Average Treatment Effect on the Treated in Mediated and Unmediated Effect",fig.align='center',out.width='50%',fig.pos='htbp'}
# plotting the result
ggplot(mediation.example, aes(x=Decomposition, y=TT,fill=Type)) +
  geom_bar(stat="identity")+
  theme_bw()
```
As Figure \@ref(fig:PlotMediationExample) shows, most of the effect of the treatment on the outcome is not mediated through $M_i$ in our example.
We can also see that there seems to be limited interaction between the treatment and the mediator since both decompositions yield similar quantities.

### Decomposing mediated and unmediated effects

In order to understand better how mediation works, it is very useful to decompose the mediated and unmediated effects using the classical typology of [Imbens and Angrist (1994)](https://www.jstor.org/stable/2951620).
Let's define four types by the reactions of their mediators to the treatment:

  - **Always takers**, who always have the mediator set to 1: $M_i^{1}=M_i^{0}=1$.
  I denote them $T_i=a$.
  - **Never takers**, who always have the mediator set to 0: $M_i^{1}=M_i^{0}=0$.
  I denote them $T_i=n$.
  - **Compliers**, whose value of the mediator moves from 0 to 1 when the value of the treatment variable changes from 0 to 1: $M_i^{1}=1$ and $M_i^{0}=0$.
  I denote them $T_i=c$.
  - **Defiers**, whose value of the mediator switches from 1 to 0 when the value of the treatment variable changes from 0 to 1: $M_i^{1}=0$ and $M_i^{0}=1$.
  I denote them $T_i=d$.

$T_i$ spans all the potential reactions of the mediator to the treatment, and thus $\cup_{t\in\{a,n,c,d\}}\{T_i=t\}$ is a partition of the sampling space.
It is very interesting to decompose both the full treatment effect on the treated but also the mediated and unmediated treatment effects on the treated in terms of the $T_i$ variable.
Let's go.
The Treatment on the Treated parameter first:

\begin{align*}
  \Delta^Y_{TT} & = \esp{Y_i^{1,M_i^1}-Y_i^{0,M_i^0}|D_i=1}\\
                & = \sum_{t\in\{a,n,c,d\}}\esp{Y_i^{1,M_i^1}-Y_i^{0,M_i^0}|T_i=t,D_i=1}\Pr(T_i=t|D_i=1)\\
                & = \esp{Y_i^{1,1}-Y_i^{0,1}|T_i=a,D_i=1}\Pr(T_i=a|D_i=1)\\
                & \phantom{=}+\esp{Y_i^{1,0}-Y_i^{0,0}|T_i=n,D_i=1}\Pr(T_i=n|D_i=1)\\
                & \phantom{=}+\esp{Y_i^{1,1}-Y_i^{0,0}|T_i=c,D_i=1}\Pr(T_i=c|D_i=1)\\
                & \phantom{=}+\esp{Y_i^{1,0}-Y_i^{0,1}|T_i=d,D_i=1}\Pr(T_i=d|D_i=1).
\end{align*}

Let us now look at the average mediated effect on the treated:

\begin{align*}
\Delta^{Y_{m(1)}}_{TT} & = \esp{Y_i^{1,M_i^1}-Y_i^{1,M_i^0}|D_i=1}\\
                & = \sum_{t\in\{a,n,c,d\}}\esp{Y_i^{1,M_i^1}-Y_i^{1,M_i^0}|T_i=t,D_i=1}\Pr(T_i=t|D_i=1)\\
                & = \esp{Y_i^{1,1}-Y_i^{1,1}|T_i=a,D_i=1}\Pr(T_i=a|D_i=1)\\
                & \phantom{=}+\esp{Y_i^{1,0}-Y_i^{1,0}|T_i=n,D_i=1}\Pr(T_i=n|D_i=1)\\
                & \phantom{=}+\esp{Y_i^{1,1}-Y_i^{1,0}|T_i=c,D_i=1}\Pr(T_i=c|D_i=1)\\
                & \phantom{=}+\esp{Y_i^{1,0}-Y_i^{1,1}|T_i=d,D_i=1}\Pr(T_i=d|D_i=1)\\
                & = \esp{Y_i^{1,1}-Y_i^{1,0}|T_i=c,D_i=1}\Pr(T_i=c|D_i=1)\\
                & \phantom{=}-\esp{Y_i^{1,1}-Y_i^{1,0}|T_i=d,D_i=1}\Pr(T_i=d|D_i=1).
\end{align*}

The result above is very useful and provides several fruitful insights.
First, the average mediated effect on the treated only involves the effects on the compliers and the defiers.
That is, only units whose value of the mediator is affected by the treatment are part of the mediated treatment effect.
Second, their importance in the mediated treatment effect depends on their proportion in the (treated) population.
That is, if few units have their value of the mediator affected by the treatment, the size of the mediated treatment effect decreases.
Finally, the impact of the mediator on the defiers enters with a negative weight in the definition of the mediated treatment effect.
This makes sense, since the treatment makes these units experience lower values of the mediator.

Finally, let us decompose the unmediated treatment effect on the treated:

\begin{align*}
\Delta^{Y_{u(0)}}_{TT} & = \esp{Y_i^{1,M_i^0}-Y_i^{0,M_i^0}|D_i=1}\\
                & = \sum_{t\in\{a,n,c,d\}}\esp{Y_i^{1,M_i^0}-Y_i^{0,M_i^0}|T_i=t,D_i=1}\Pr(T_i=t|D_i=1)\\
                & = \esp{Y_i^{1,1}-Y_i^{0,1}|T_i=a,D_i=1}\Pr(T_i=a|D_i=1)\\
                & \phantom{=}+\esp{Y_i^{1,0}-Y_i^{0,0}|T_i=n,D_i=1}\Pr(T_i=n|D_i=1)\\
                & \phantom{=}+\esp{Y_i^{1,0}-Y_i^{0,0}|T_i=c,D_i=1}\Pr(T_i=c|D_i=1)\\
                & \phantom{=}+\esp{Y_i^{1,1}-Y_i^{0,1}|T_i=d,D_i=1}\Pr(T_i=d|D_i=1).
\end{align*}

The unmediated treatment effect on the treated is equal to the direct effect of the treatment on each of the types multiplied by their proportion in the (treated) population, keeping the value of the mediator at its default value when the treatment is equal to 0.

## The Fundamental Problem of Mediation Analysis

In this section, we state the Fundamental Problem of Mediation Analysis (FPMA) and we then move on to examining the biases of intuitive comparisons that could be used to recover the mediated and unmediated effects of the treatment.

### The Fundamental Problem of Mediation Analysis

Mediation analysis is hard.
It is even harder than identification and estimation of treatment effects.
The following theorem shows formally why this is the case:

```{theorem,FPMA,name="Fundamental Problem of Mediation Analysis"}
The mediated and unmediated treatment effects are fundamentally unobservable.
```

```{proof}
Let's look at the mediated treatment effect first.
It can be decomposed as follows:

\begin{align*}
  \Delta^{Y_{m(1)}}_{TT} & = \esp{Y_i^{1,M_i^1}-Y_i^{1,M_i^0}|D_i=1}\\
                        &  = \esp{Y_i^{1,M_i^1}-Y_i^{1,M_i^0}|D_i=1,M_i=1}\Pr(M_i=1|D_i=1)\\
                        & \phantom{=} + \esp{Y_i^{1,M_i^1}-Y_i^{1,M_i^0}|D_i=1,M_i=0}\Pr(M_i=0|D_i=1)\\
                        & = \esp{Y^{1,1}_i-Y_i^{1,M_i^0}|D_i=1,M_i=1}\Pr(M_i=1|D_i=1)\\
                        & \phantom{=} + \esp{Y^{1,0}_i-Y_i^{1,M_i^0}|D_i=1,M_i=0}\Pr(M_i=0|D_i=1)\\
                        & = (\esp{Y_i|D_i=1,M_i=1}-\esp{Y_i^{1,M_i^0}|D_i=1,M_i=1})\Pr(M_i=1|D_i=1)\\
                        & \phantom{=} + (\esp{Y_i|D_i=1,M_i=0}-\esp{Y_i^{1,M_i^0}|D_i=1,M_i=0})\Pr(M_i=0|D_i=1),
\end{align*}

where the second and third inequalities follow from the switching equation.
Note that both $\esp{Y_i^{1,M_i^0}|D_i=1,M_i=1}$ and $\esp{Y_i^{1,M_i^0}|D_i=1,M_i=0}$ are counterfactual values that cannot be observed (they depend on the value of the mediator when $D_i=0$, which is unobserved when $D_i=1$).

Let us look at the unmediated treatment effect now:

\begin{align*}
  \Delta^{Y_{u(1)}}_{TT} & = \esp{Y_i^{1,M_i^1}-Y_i^{0,M_i^1}|D_i=1}\\
                         &  = \esp{Y_i^{1,M_i^1}-Y_i^{0,M_i^1}|D_i=1,M_i=1}\Pr(M_i=1|D_i=1)\\
                        & \phantom{=} + \esp{Y_i^{1,M_i^1}-Y_i^{0,M_i^1}|D_i=1,M_i=0}\Pr(M_i=0|D_i=1)\\
                        & = \esp{Y^{1,1}_i-Y_i^{0,M_i^1}|D_i=1,M_i=1}\Pr(M_i=1|D_i=1)\\
                        & \phantom{=} + \esp{Y^{1,0}_i-Y_i^{0,M_i^1}|D_i=1,M_i=0}\Pr(M_i=0|D_i=1)\\
                        & = (\esp{Y_i|D_i=1,M_i=1}-\esp{Y_i^{0,M_i^1}|D_i=1,M_i=1})\Pr(M_i=1|D_i=1)\\
                        & \phantom{=} + (\esp{Y_i|D_i=1,M_i=0}-\esp{Y_i^{0,M_i^1}|D_i=1,M_i=0})\Pr(M_i=0|D_i=1),
\end{align*}

Note that both $\esp{Y_i^{0,M_i^1}|D_i=1,M_i=1}$ and $\esp{Y_i^{0,M_i^1}|D_i=1,M_i=0}$ are counterfactual values that cannot be observed (they depend on the value of the outcome when $D_i=0$, which is unobserved when $D_i=1$).

```

```{remark}
The key to understanding Theorem \@ref(thm:FPMA) is to see that the potential value of the mediator is unknown and this combines with the fact that the potential value of the outcome had the mediator or the treatment changed value is also unobserved.
```

### Biases of Intuitive Comparisons

Usual intuitive proxies for the mediated and unmediated effects are usually biased because of confounders.
We can visualize the problem in a DAG, with the arcs indicating the existence of confounders between two variables:

```{r DAGMediationConfounded,eval=TRUE,echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,results='hide',fig.cap='DAG of a confounded mediating relationship',fig.align='center',out.width='50%',fig.pos='htbp'}
# structure of DAG
our_dag <- ggdag::dagify(M ~ D,
                         Y ~ D,
                         Y ~ M,
                         M ~~ D,
                         Y ~~ D,
                         Y ~~ M,
                         coords=list(x = c(D = 0, M = 1, Y = 2),
                                     y = c(D = 0, M = 1, Y = 0))
                        )
# plotting DAG
ggdag::ggdag(our_dag) + 
  theme_void()
```
The most obvious intuitive comparison that could try to recover mediated and unmediated treatment effects is the conditional With/Without comparison:

\begin{align*}
  \Delta^Y_{WW|D_i=d} & = \esp{Y_i|D_i=d,M_i=1}-\esp{Y_i|D_i=d,M_i=0}\\
  \Delta^Y_{WW|M_i=m} & = \esp{Y_i|D_i=1,M_i=m}-\esp{Y_i|D_i=0,M_i=m}.
\end{align*}

Our hope is that $\Delta^Y_{WW|D_i=d}$ recovers the mediated treatment effect and $\Delta^Y_{WW|M_i=d}$ recovers the unmediated treatment effect.
Unfortunately, that is not generally the case.
The following theorem makes this point clear:

```{theorem,WWBiasMed,name="Bias of the Conditional With/Without Comparisons"}
The conditional With/Without comparisons are biased estimators of the mediated and unmediated average effects of the treatment on the treated:
  
\begin{align*}
  \Delta^Y_{WW|D_i=1} & = \Delta^{Y_{m(1)}}_{TT|M_i=1}+\Delta^{Y^{1,0}}_{SB^{m(1)}|D_i=1}\\
  \Delta^Y_{WW|M_i=1} & = \Delta^{Y_{u(1)}}_{TT|M_i=1}+\Delta^{Y^{0,1}}_{SB^{u(1)}|M_i=1}.
\end{align*}

with 

\begin{align*}
  \Delta^{Y_{m(1)}}_{TT|M_i=1} & = \esp{Y_i^{1,1}-Y_i^{1,0}|D_i=1,M_i=1}\\
  \Delta^{Y^{1,0}}_{SB^{m(1)}|D_i=1} & = \esp{Y^{1,0}_i|D_i=1,M_i=1}-\esp{Y^{1,0}_i|D_i=1,M_i=0}\\
  \Delta^{Y_{u(1)}}_{TT|M_i=1} & = \esp{Y_i^{1,1}-Y_i^{0,1}|D_i=1,M_i=1}\\
  \Delta^{Y^{0,1}}_{SB^{u(1)}|M_i=1} & = \esp{Y^{0,1}_i|D_i=1,M_i=1}-\esp{Y^{0,1}_i|D_i=0,M_i=0}.
\end{align*}

```

```{proof}
The proof is straightforward using the switching equation and adding and subtracting the counterfactual part of the treatment effects.

```

According to Theorem \@ref(thm:WWBiasMed), Selection Bias biases the conditional With/Without estimators of the mediated and unmediated treatment effects.
As Figure \@(ref:DAGMediationConfounded) suggests, this is because confounders are correlated with both treatment, mediator and outcome.
The most important confounder in our example is $\mu_i$, which is correlated with all three variables.

```{remark}
Note that Theorem \@ref(thm:WWBiasMed) shows the bias for the mediated and unmediated treatment effects conditional on $M_i=1$. 
The same results with the unconditional version of the treatment effects would be somewhat more intricate since it would involve weights.
```

```{remark}
Note also that Theorem \@ref(thm:WWBiasMed) shows the bias for the mediated and unmediated treatment effects with $m(1)$ and $u(1)$ as indices. 
The same results with $m(0)$ and $u(0)$ are similar.
```

```{example}
Let's see how that works our in our example.
```

```{r BiasWWMed,eval=TRUE,echo=TRUE,results='hide'}
# mediated treatment effect for M=1
TTm1M1 <- mean(y11[Ds==1&M==1])-mean(y10[Ds==1&M==1])
# unmediated treatment effect for M==1
TTu1M1 <- mean(y11[Ds==1&M==1])-mean(y01[Ds==1&M==1])

# selection bias for the mediated treatment effect
SBm1D1 <- mean(y10[Ds==1&M==1])-mean(y10[Ds==1&M==0])
# selection bias for the unmediated treatment effect
SBu1M1 <- mean(y01[Ds==1&M==1])-mean(y01[Ds==0&M==1])

# with/without comparison for the mediated treatment effect for M=1
WWD1 <- mean(y[Ds==1&M==1])-mean(y[Ds==1&M==0])
# unmediated treatment effect for M==1
WWM1 <- mean(y[Ds==1&M==1])-mean(y[Ds==0&M==1])
```

In our example, the With/Without estimator of the mediated treatment effect is equal to `r round(WWD1,3)` (for $D_i=1$).
The true mediated treatment effect for $D_i=1$ and $M_i=1$ is equal to `r round(TTm1M1,3)`.
Selection bias for the mediated treatment effect for $D_i=1$ is thus equal to `r round(SBm1D1,3)`.
The With/Without estimator of the unmediated treatment effect is equal to `r round(WWM1,3)` (for $M_i=1$).
The true unmediated treatment effect for $D_i=1$ and $M_i=1$ is equal to `r round(TTu1M1,3)`.
Selection bias for the unmediated treatment effect for $M_i=1$ is thus equal to `r round(SBu1M1,3)`.

## Mediation analysis with experimental data

[Imai, Tingleu and Yamamoto (2013)](https://onlinelibrary.wiley.com/doi/abs/10.1111/j.1467-985X.2012.01032.x) propose two experimental designs to recover the mediated and unmediated treatment effects: the parallel design and the crossover design.
We are going to also study identification in a Sequential Self-Selection design, which I find slightly simpler. 
Let's examine these approaches in turn.

### Mediation analysis in the Parallel design

In the parallel design, we run a three-step experiment.
In the first step, we randomly allocate units either to the first or the second arm of the experiment.
We denote $R^1_i=0$ the first arm and $R^1_i=1$ the second arm.

In the first arm, we randomly allocate the treatment among units and we let the mediator take the corresponding value without interferring.
We denote $R^2_i=1$ when unit $i$ is assigned to the treatment and $R^2_i=0$ when unit $i$ us not assigned to the treatment. 
In this design, we assume that units fully comply with the assignment.
One way to do that is to run the whole parallel design only on units that are selected into the treatment ($D_i=1$).

In the second arm, we run a two-step experiment.
In the first step, we randomly allocate units either to the treatment or to the control.
We denote $R_i^2=1$ when unit $i$ is assigned to the treatment and $R_i^2=0$ when it is assigned to the control.
In the second step, we randomly allocate units to the mediator. 
We denote $R_i^3=1$ when unit $i$ is assigned to the mediator value $M_i=1$ and $R_i^3=0$ when unit $i$ is assigned to the mediator value of $M_i=0$.

In order to recover the treatment effect on the treated, we are going to assume that the experiment takes place among the units selected to participate in the treatment, *i.e.* among units with $D_i=1$.
If we abstract from this, the parameter we recover is the average effect of the treatment, ATE.

Let us formalize the assumptions we make on the Parallel design

```{hypothesis,parindepss,name='Sequential Independence in the Parallel Design'}
We assume that the sequential randomized allocation of the program among applicants is well done: 

\begin{align*}
  R^1_i & \Ind(Y_i^{00},Y_i^{01},Y_i^{10},Y_i^{11},M^1_i,M^0_i)|D_i=1\\
  R^2_i & \Ind(Y_i^{00},Y_i^{01},Y_i^{10},Y_i^{11},M^1_i,M^0_i)|R_i^1,D_i=1\\
  R^3_i & \Ind(Y_i^{00},Y_i^{01},Y_i^{10},Y_i^{11},M^1_i,M^0_i)|R_i^2,R_i^1=1,D_i=1.
\end{align*}

```

We need a second assumption:

```{hypothesis,parrassval,name='Validity of the Sequential Parallel Design'}
We assume that the sequential randomized allocation of the program does not interfere with how potential outcomes and potential treatment and mediator choices are generated:

\begin{align*}
Y_i & = 
  \begin{cases}
    Y_i^{11} & \text{ if } D_i=1 \text{ and } ((R^1_i=0 \text{ and } R^2_i=1 \text{ and } M_i=1) \text{ or } (R^1_i=1 \text{ and } R^2_i=1 \text{ and } R^3_i=1))   \\
    Y_i^{10} & \text{ if } D_i=1 \text{ and } ((R^1_i=0 \text{ and } R^2_i=1 \text{ and } M_i=0) \text{ or } (R^1_i=1 \text{ and } R^2_i=1 \text{ and } R^3_i=0))   \\
    Y_i^{01} & \text{ if } D_i=1 \text{ and } ((R^1_i=0 \text{ and } R^2_i=0 \text{ and } M_i=1) \text{ or } (R^1_i=1 \text{ and } R^2_i=0 \text{ and } R^3_i=1))   \\
    Y_i^{00} & \text{ if } D_i=1 \text{ and } ((R^1_i=0 \text{ and } R^2_i=0 \text{ and } M_i=0) \text{ or } (R^1_i=1 \text{ and } R^2_i=0 \text{ and } R^3_i=0))   \\
  \end{cases}\\
M_i & = 
  \begin{cases}
    M_i^{1} & \text{ if } D_i=1 \text{ and }R^1_i=0\text{ and } R^2_i=1\\
    M_i^{0} & \text{ if } D_i=1 \text{ and }R^1_i=0\text{ and } R^2_i=0
  \end{cases},
\end{align*}

with $(Y_i^{00},Y_i^{01},Y_i^{10},Y_i^{11},M^1_i,M^0_i)$ the same potential outcomes and potential treatment and mediator choices as in a routine allocation of the treatment.
```

```{remark}
Assumption \@ref(hyp:parrassval) imposes that the experimental setting does not alter the treatment, mediator and outcomes of the agents.
```

For identification of the mediated and unmediated effects to go through, we need a final (strong) assumption: the absence of interaction effects:

```{hypothesis,ParNoInteract,name='Absence of Interaction Effects'}
We assume that the treatment and the mediator do not interact: $Y_i^{d,m}-Y_i^{d,m'}=Y_i^{d',m}-Y_i^{d',m'}$, $\forall (d,m,d'm')\in\{0,1\}^4$, $d\neq d'$, $m\neq m'$.
```

```{remark}
Assumption \@ref(hyp:ParNoInteract) is actually too strong for our result.
We only need that the assumption holds for the average effects on the *never takers*, *defiers* and *compliers*.
```

```{remark}
Under Assumption \@ref(hyp:ParNoInteract), $\Delta^{Y_{u(1)}}_{TT}=\Delta^{Y_{u(0)}}_{TT}$, which we denote $\Delta^{Y_{u(d)}}_{TT}$ and $\Delta^{Y_{m(1)}}_{TT}=\Delta^{Y_{m(0)}}_{TT}$, which we denote $\Delta^{Y_{m(d)}}_{TT}$.
```

Under these assumptions, we have the following result:

```{theorem,idParMed,name='Identification of Total, Mediated and Unmediated Effects in the Parallel Design'}
Under Assumptions \@ref(hyp:parindepss), \@ref(hyp:parrassval) and \@ref(hyp:ParNoInteract), the treatment effect on the treated and the mediated and unmediated effects on the treated are identified in the Parallel Design:

\begin{align*}
  \Delta^Y_{TT} & = \esp{Y_i|D_i=1,R^1_i=0,R_i^2=1}-\esp{Y_i|D_i=1,R^1_i=0,R_i^2=0}\\
  \Delta^{Y_{u(d)}}_{TT} & = \esp{Y_i|D_i=1,R^1_i=1,R_i^2=1,R_i^3=1}-\esp{Y_i|D_i=1,R^1_i=1,R_i^2=0,R_i^3=1}\\
                        & =\esp{Y_i|D_i=1,R^1_i=1,R_i^2=1,R_i^3=0}-\esp{Y_i|D_i=1,R^1_i=1,R_i^2=0,R_i^3=0}\\
  \Delta^{Y_{m(d)}}_{TT} & = \esp{Y_i|D_i=1,R^1_i=0,R_i^2=1}-\esp{Y_i|D_i=1,R^1_i=0,R_i^2=0}\\
                        & \phantom{=}-\esp{Y_i|D_i=1,R^1_i=1,R_i^2=1,R_i^3=1}-\esp{Y_i|D_i=1,R^1_i=1,R_i^2=0,R_i^3=1}\\
                         & = \esp{Y_i|D_i=1,R^1_i=0,R_i^2=1}-\esp{Y_i|D_i=1,R^1_i=0,R_i^2=0}\\
                        & \phantom{=}-\esp{Y_i|D_i=1,R^1_i=1,R_i^2=1,R_i^3=0}-\esp{Y_i|D_i=1,R^1_i=1,R_i^2=0,R_i^3=0}.
\end{align*}
```

```{proof}
I omit conditioning on $D_i=1$ to save notation.
Let us start with the total effect: $\esp{Y_i|R^1_i=0,R_i^2=1}-\esp{Y_i|R^1_i=0,R_i^2=0}$ is a With/Without estimator where the treatment has been randomized among a random sample of the units selected to receive the treatment. 
Theorem \@ref(thm:idrass) proves that it identifies $\Delta^Y_{TT}$ under Assumptions \@ref(hyp:parindepss) and \@ref(hyp:parrassval).
The unmediated treatment effect can be recovered as follows:

\begin{align*}
\esp{Y_i|R^1_i=1,R_i^2=1,R_i^3=1} & -\esp{Y_i|R^1_i=1,R_i^2=0,R_i^3=1} \\
                        & = \sum_{t\in\{a,n,c,d\}}\esp{Y_i^{1,1}-Y_i^{0,1}|T_i=t}\Pr(T_i=t)\\
\esp{Y_i|R^1_i=1,R_i^2=1,R_i^3=0} & -\esp{Y_i|R^1_i=1,R_i^2=0,R_i^3=0} \\
                        & = \sum_{t\in\{a,n,c,d\}}\esp{Y_i^{1,0}-Y_i^{0,0}|T_i=t}\Pr(T_i=t),
\end{align*}

using Assumptions \@ref(hyp:parindepss) and \@ref(hyp:parrassval).
Using the decomposition of $\Delta^Y_{TT}$ in the effects on each type, we have:

\begin{align*}
\esp{Y_i|R^1_i=0,R_i^2=1} & -\esp{Y_i|R^1_i=0,R_i^2=0}\\
  & - (\esp{Y_i|R^1_i=1,R_i^2=1,R_i^3=1} -\esp{Y_i|R^1_i=1,R_i^2=0,R_i^3=1}) \\
  & = \esp{Y_i^{0,1}-Y_i^{0,0}|T_i=c}\Pr(T_i=c) \\
  & \phantom{=}-\esp{Y_i^{1,1}-Y_i^{1,0}|T_i=d}\Pr(T_i=d)\\
  & \phantom{=}+(\esp{Y_i^{1,0}-Y_i^{0,0}|T_i=n}-\esp{Y_i^{1,1}-Y_i^{0,1}|T_i=n})\Pr(T_i=n).
\end{align*}

and 

\begin{align*}
\esp{Y_i|R^1_i=0,R_i^2=1} & -\esp{Y_i|R^1_i=0,R_i^2=0}\\
  & - (\esp{Y_i|R^1_i=1,R_i^2=1,R_i^3=0} -\esp{Y_i|R^1_i=1,R_i^2=0,R_i^3=0}) \\
  & = \esp{Y_i^{1,1}-Y_i^{1,0}|T_i=c}\Pr(T_i=c) \\
  & \phantom{=}-\esp{Y_i^{0,1}-Y_i^{0,0}|T_i=d}\Pr(T_i=d)\\
  & \phantom{=}+(\esp{Y_i^{1,1}-Y_i^{0,1}|T_i=a}-\esp{Y_i^{1,0}-Y_i^{0,0}|T_i=a})\Pr(T_i=a).
\end{align*}

Using Assumption \@ref(hyp:ParNoInteract) and the decomposition formulae for the mediated and unmediated treatment effects on the treated gives the result.

```

```{remark}
The proof of Theorem \@ref(thm:idParMed) is more detailed and explanatory than the one in [Imai, Tingleu and Yamamoto (2013)](https://onlinelibrary.wiley.com/doi/abs/10.1111/j.1467-985X.2012.01032.x).
We understand why we need Assumption \@ref(hyp:ParNoInteract): the effects on the *always takers* and on the *never takers* estimated in the two treatment arms do not cancel out and the effects of the mediator on the *compliers* and *defiers* are not the correct combination in both comparisons unless we make Assumption \@ref(hyp:ParNoInteract).
```

```{remark}
Note that the model is overidentified under Assumptions \@ref(hyp:parindepss), \@ref(hyp:parrassval) and \@ref(hyp:ParNoInteract).
There are two different ways to estimate the unmediated treatment effect and thus one can test Assumption \@ref(hyp:ParNoInteract) by comparing these two estimates.
```

```{remark}
Since the model is overidentified under Assumptions \@ref(hyp:parindepss), \@ref(hyp:parrassval) and \@ref(hyp:ParNoInteract), one wonders what can exactly be identified on the mediated and unmediated treatment effects under these assumptions.
Maybe  it is possible to bound the treatment effects.
This is, to my knowledge, still an open question.
```

### Mediation analysis in the Sequential Self-Selection design

In a Sequential Self-Selection design, researchers run a Randomized Controlled Trial (RCT) in two-steps.
In the first step, the treatment is randomized among those who applied for it.
In the second step, the mediator is randomized among those who applied for it. 
We thus have two random variables $R_i^1$ and $R_i^2$ which take values zero or one for observations for which $D_i=1$ for the first and $D_i=1$ and $M_i=1$ for the second.

Let us formalize how the Sequential Self-Selection design works.

```{hypothesis,seqindepss,name='Sequential Independence Among Self-Selected'}
We assume that the sequential randomized allocation of the program among applicants is well done: 

\begin{align*}
  R^1_i & \Ind(Y_i^{00},Y_i^{01},Y_i^{10},Y_i^{11},M^1_i,M^0_i)|D_i=1\\
  R^2_i & \Ind(Y_i^{00},Y_i^{01},Y_i^{10},Y_i^{11},M^1_i,M^0_i)|R^1_i,M_i=1,D_i=1
\end{align*}

```

We need a second assumption:

```{hypothesis,seqrassval,name='Validity of the Sequential Self-Selection design'}
We assume that the sequential randomized allocation of the program does not interfere with how potential outcomes and self-selection are generated:

\begin{align*}
Y_i & = 
  \begin{cases}
    Y_i^{11} & \text{ if } (D_i=1 \text{ and } R^1_i=1 \text{ and } M_i=1 \text{ and } R^2_i=1)   \\
    Y_i^{10} & \text{ if } ((D_i=1 \text{ and } R^1_i=1) \text{ and } ((M_i=1 \text{ and } R^2_i=0) \text{ or } M_i=0))    \\
    Y_i^{01} & \text{ if } (D_i=1 \text{ and } R^1_i=0 \text{ and } M_i=1 \text{ and } R^2_i=1)   \\
    Y_i^{00} & \text{ if } ((D_i=1 \text{ and } R^1_i=0) \text{ and } ((M_i=1 \text{ and } R^2_i=0) \text{ or } M_i=0))   
  \end{cases}\\
M_i & = 
  \begin{cases}
    M_i^{1} & \text{ if } D_i=1 \text{ and } R^1_i=1\\
    M_i^{0} & \text{ if } D_i=1 \text{ and } R^1_i=0
  \end{cases}
\end{align*}
```

with $(Y_i^{00},Y_i^{01},Y_i^{10},Y_i^{11},M^1_i,M^0_i)$ the same potential outcomes and self-selection decisions as in a routine allocation of the treatment.

We finally need a monotonicity assumption:

```{hypothesis,seqMono,name='Monotonicity of the Effect of the Treatment on the Mediator'}
We assume that the effect of the treatment on the mediator is monotonous:

\begin{align*}
  \Pr(M_i^1-M_i^0<0) & =0.
\end{align*}
```

```{remark}
Clearly, Assumption \@ref(hyp:seqMono) is equivalent to the absence of *defiers*: $\Pr(T_i=d)=0$.
```

Under these assumptions, we have the following result:

```{theorem,idrassMed,name='Identification of Total, Mediated and Unmediated Effects in a Sequential Self-Selection design'}
Under Assumptions \@ref(hyp:seqindepss), \@ref(hyp:seqrassval) and \@ref(hyp:seqMono), the treatment effect on the treated and the mediated and unmediated effects are identified in a Sequential Self-Selection Design:

\begin{align*}
  \Delta^Y_{TT} & = \esp{Y_i|D_i=1,R_i^1=1,M_i=1,R_i^2=1}\Pr(M_i=1|D_i=1,R_i^1=1)\\
                & \phantom{=}+\esp{Y_i|D_i=1,R_i^1=1,M_i=0}\Pr(M_i=0|D_i=1,R_i^1=1)\\
                & \phantom{=}-\left(\esp{Y_i|D_i=1,R_i^1=0,M_i=1,R_i^2=1}\Pr(M_i=1|D_i=1,R_i^1=0)\right.\\
                & \phantom{=-(}\left.+\esp{Y_i|D_i=1,R_i^1=0,M_i=0}\Pr(M_i=0|D_i=1,R_i^1=0)\right)\\
   \Delta^{Y_{m(1)}}_{TT} & =\left(\esp{Y_i|D_i=1,R_i^1=1,M_i=1,R_i^2=1}\right.\\
                          &\phantom{=-(}\left.-\esp{Y_i|D_i=1,R_i^1=1,M_i=1,R_i^2=0}\right)\Pr(M_i=1|D_i=1,R_i^1=1)\\
                          & \phantom{=}-\left(\esp{Y_i|D_i=1,R_i^1=0,M_i=1,R_i^2=1}\right.\\
                          &\phantom{=-(}\left.-\esp{Y_i|D_i=1,R_i^1=0,M_i=1,R_i^2=0}\right)\Pr(M_i=1|D_i=1,R_i^1=0)\\
   \Delta^{Y_{u(0)}}_{TT} & =\Delta^Y_{TT}-\Delta^{Y_{m(1)}}_{TT}.
\end{align*}

```

```{proof}
I omit conditioning on $D_i=1$ all along to save on notation.
Note that:

\begin{align*}  
  \esp{Y_i|R_i^1=1,M_i=1,R_i^2=1} & = \esp{Y^{1,M_i^1}_i|R_i^1=1,M^1_i=1,R_i^2=1}\\
                                  & = \esp{Y^{1,M_i^1}_i|M^1_i=1}\\
  \esp{Y_i|R_i^1=1,M_i=0} & = \esp{Y^{1,M_i^1}_i|M^1_i=0}\\
  \esp{Y_i|R_i^1=0,M_i=1,R_i^2=1} & = \esp{Y^{0,M_i^0}_i|M^0_i=1}\\
  \esp{Y_i|R_i^1=0,M_i=0} & = \esp{Y^{0,M_i^0}_i|M^0_i=0}\\
  \Pr(M_i=1|R_i^1=1) & = \Pr(M^1_i=1)\\
  \Pr(M_i=0|R_i^1=1) & = \Pr(M^1_i=0)\\
  \Pr(M_i=1|R_i^1=0) & = \Pr(M^0_i=1)\\
  \Pr(M_i=0|R_i^1=0) & = \Pr(M^0_i=0),
\end{align*}

using Assumptions \@ref(hyp:seqindepss) and \@ref(hyp:seqrassval).
Because $M_i^1$ and $M_i^0$ are both partitions of the sampling space, we can use the Law of Iterated Expectations to find:
  
\begin{align*}
  \esp{Y^{1,M_i^1}_i} &=  \esp{Y^{1,M_i^1}_i|M^1_i=1}\Pr(M^1_i=1)+\esp{Y^{1,M_i^1}_i|M^1_i=1}\Pr(M^1_i=0)\\
  \esp{Y^{0,M_i^0}_i} &=  \esp{Y^{0,M_i^0}_i|M^0_i=1}\Pr(M^0_i=1)+\esp{Y^{0,M_i^0}_i|M^1_i=1}\Pr(M^0_i=0)
\end{align*}
  
The difference between the two terms is the definition of $\Delta^Y_{TT}$.
This proves the first part of the theorem.

Note that, under Assumptions \@ref(hyp:seqindepss), \@ref(hyp:seqrassval) and especially \@ref(hyp:seqMono):

\begin{align*}
  \esp{Y_i|R_i^1=1,M_i=1,R_i^2=1} & -\esp{Y_i|R_i^1=1,M_i=1,R_i^2=0}\\
     & =  \esp{Y^{11}_i|M^1_i=1} -\esp{Y^{10}_i|M^1_i=1}\\
     & =  \esp{Y^{11}_i|T_i=a}\Pr(T_i=a|M_i^1=1)+\esp{Y^{11}_i|T_i=c}\Pr(T_i=c|M_i^1=1)\\
      & \phantom{=}-\esp{Y^{10}_i|T_i=a}\Pr(T_i=a|M^1_i=1)-\esp{Y^{10}_i|T_i=c}\Pr(T_i=c|M^1_i=1)\\
     & =  \esp{Y^{11}_i-Y^{10}_i|T_i=a}\Pr(T_i=a|M_i^1=1)\\
      & \phantom{=}+\esp{Y^{11}_i-Y^{10}_i|T_i=c}\Pr(T_i=c|M_i^1=1).
\end{align*}

By the same reasoning, we also have:
  
\begin{align*}
  \esp{Y_i|R_i^1=0,M_i=1,R_i^2=1} & -\esp{Y_i|R_i^1=0,M_i=1,R_i^2=0}\\
     & =  \esp{Y^{11}_i|M^0_i=1} -\esp{Y^{10}_i|M^0_i=1}\\
     & =  \esp{Y^{11}_i-Y^{10}_i|T_i=a}.
\end{align*}
  
Finally, using Bayes Law and the fact that:
  
\begin{align*}
  \Pr(M_i=1|R_i^1=1) & = \Pr(M^1_i=1) = \Pr(T_i=a)+\Pr(T_i=c)\\
  \Pr(M_i=1|R_i^1=0) & = \Pr(M^0_i=1) = \Pr(T_i=a),
\end{align*}

we have that $\Pr(T_i=a|M_i^1=1)=\frac{\Pr(T_i=a)}{\Pr(T_i=a)+\Pr(T_i=c)}$ and  $\Pr(T_i=c|M_i^1=1)=\frac{\Pr(T_i=c)}{\Pr(T_i=a)+\Pr(T_i=c)}$.
Combining the results with the decomposition of $\Delta^{Y_{m(1)}}_{TT}$ under Assumption \@ref(hyp:seqMono) proves the identification of $\Delta^{Y_{m(1)}}_{TT}$. 

The identification of $\Delta^{Y_{u(0)}}_{TT}$ comes directly from the decompositions of $\Delta^Y_{TT}$, $\Delta^{Y_{m(1)}}_{TT}$ and $\Delta^{Y_{u(0)}}_{TT}$ under Assumption \@ref(hyp:seqMono).
This proves the result.

```




## Mediation analysis under unconfoundedness

## Mediation analysis with panel data

## Mediation analysis with instruments
